---
title: "make dataset"
author: "doh"
date: "`r Sys.Date()`"
output: html_document
---

 
# BP and UPF and Na in NDNS Dissertation calculation and results

```{r include=FALSE}
library(haven)
library(data.table)
library(readxl)
library(sensemakr)
library(labelled)
library(survey)
```


```{r read data, message=FALSE, warning=FALSE, include=FALSE}

sav4 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_personleveldietarydata_uk_v2.sav"

sav4b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_indiv_uk.sav"

sav6 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_personleveldietarydata_v2.sav"

sav6b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_indiv.sav"

sav8 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_personleveldietarydata.sav"

sav8b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_indiv.sav"



sav11 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_personleveldietarydata_uk_20210831.sav"

sav11b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_indiv_20211020.sav"

sav4r <- as.data.table(read_sav(sav4))
sav4br <-as.data.table(read_sav(sav4b))

sav6r <- as.data.table(read_sav(sav6))
sav6br <-as.data.table(read_sav(sav6b))

sav8r <- as.data.table(read_sav(sav8))
sav8br <-as.data.table(read_sav(sav8b))


sav11r <- as.data.table(read_sav(sav11))
sav11br <- as.data.table(read_sav(sav11b))
sav11br$area <- sav11br$Area
```


```{r std names4, message=FALSE, warning=FALSE, include=FALSE}
#standardise the variable names for grp 1-4 
sav4br$bpmedc2 <- sav4br$bpmedc
sav4br$bpmedd2 <- sav4br$bpmedd
sav4br$diur2 <- sav4br$diur
sav4br$beta2 <- sav4br$beta
sav4br$aceinh2 <- sav4br$aceinh
sav4br$calciumb2 <- sav4br$calciumb
sav4br$obpdrug2 <- sav4br$obpdrug

sav4br$ethgrp5 <- sav4br$ethgr5
sav4br$ethgrp2 <- sav4br$ethgr2

sav4br$hyper140_2 <- sav4br$hyper140
sav4br$hibp140_2 <- sav4br$hibp140
sav4br$hyper1_2 <- sav4br$hyper1
sav4br$highbp1_2 <- sav4br$highbp1

sav4br$educfinh <- sav4br$EducFin
sav6r$SurveyYear <- sav6r$Surveyyear
```

```{r stdnames11, message=FALSE, warning=FALSE, include=FALSE}
sav11br$htval <- sav11br$htval2
sav11br$wtval <- sav11br$wtval2
sav11br$bmival <- sav11br$bmival2

sav11br$gor <- sav11br$GOR
sav11br$age <- sav11br$AgeR
sav11r$Age <- sav11r$AgeR
```


```{r novafile, message=FALSE, warning=FALSE, include=FALSE }
NOVA_ForDavid <- as.data.table(read_excel("data/NOVA_ForDavid.xlsx"))

NOVA_ForDavid[, Sex := factor(Sex, levels = 1:2)]
NOVA_ForDavid[, Country := factor(Country)]
NOVA_ForDavid[, SurveyYear := factor(SurveyYear)]
NOVA_ForDavid[, Age := as.numeric(Age)]
```




```{r weight, echo=FALSE}
#Doing the weighting correction, as explain in the PDF "6533_ndns_yrs9-11_weights_guide"

n1 = sum(sav4br$wti_UKY1234)  #Should be 6828

n2 = sum(sav6br$wti_Y56)         #Should be 2546

n3 = sum(sav8br$wti_Y78)         #Should be 2723

n4 = sum(sav11br$wti_Y911)       #Should be 3558

N <- sum(n1, n2, n3, n4) #Should be 15655
```


now the proportion weight for each is calculated

```{r}
sav4br$wti_UKY1to11 <- ((sav4br$wti_UKY1234/n1) * N)  * (4/11)

sav6br$wti_UKY1to11 <- ((sav6br$wti_Y56/n2) * N) * (2/11)

sav8br$wti_UKY1to11 <- ((sav8br$wti_Y78/n3) * N) * (2/11)

sav11br$wti_UKY1to11 <- ((sav11br$wti_Y911/n4) * N) * (3/11)

```


```{r weightn, include=FALSE}
#Doing the weighting correction, as explain in the PDF "6533_ndns_yrs9-11_weights_guide"

n1 = sum(sav4br$wtn_UKY1234)  #Should be 6828

n2 = sum(sav6br$wtn_Y56)         #Should be 2546

n3 = sum(sav8br$wtn_Y78)         #Should be 2723

n4 = sum(sav11br$wtn_Y911)       #Should be 3558

N <- sum(n1, n2, n3, n4) #Should be 15655
```


now the proportion weight for each is calculated

```{r include=FALSE}
sav4br$wtn_UKY1to11 <- ((sav4br$wtn_UKY1234/n1) * N)  * (4/11)

sav6br$wtn_UKY1to11 <- ((sav6br$wtn_Y56/n2) * N) * (2/11)

sav8br$wtn_UKY1to11 <- ((sav8br$wtn_Y78/n3) * N) * (2/11)

sav11br$wtn_UKY1to11 <- ((sav11br$wtn_Y911/n4) * N) * (3/11)

```

```{r allsetup variables, message=FALSE, warning=FALSE, include=FALSE}
 #measured and recorded data yr1-4a_indiv_uk 
#subsets of the table to identify grouped information
bpset <- c("seriali","omsysval","omdiaval")
ethnset <- c("ethgrp5","ethgrp2")
saltset <- c("SaltChk","SalHowC","SltSHow" ,"vegetarn")
medsset <- c( "bpmedc2","bpmedd2","diur2","beta2","aceinh2", "calciumb2","obpdrug2", "PregNowB")
hypset <- c("hyper140_2", "hibp140_2", "hyper1_2", "highbp1_2")
incset <-c("nssec8","gor","region", "EIMD_2007_quintile","EIMD_2010_quintile","EIMD_2015_quintile","educfinh")
measset <- c("htval","wtval","bmival")
ageset <- c("agegad1", "agegad2", "agegch1","agegr1","age")
weightset <- c("wti_UKY1to11","wtn_UKY1to11","astrata1","area")

indiv_data <- c(bpset,ethnset,saltset,medsset,hypset,incset,measset,ageset,weightset)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
sav11br <- sav11br[,..indiv_data]
sav4br <- sav4br[,..indiv_data]
sav6br <- sav6br[,..indiv_data]
sav8br <- sav8br[,..indiv_data]

val_labels(sav8br) <- NULL
val_labels(sav6br) <- NULL
val_labels(sav4br) <- NULL
val_labels(sav11br) <- NULL
#sav8br$seriali <- as.double(sav8br$seriali)
#sav8br$ethgrp5 <- as.double(sav8br$ethgrp5)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
#this creates the whole table
ndns_1_11 <- rbindlist(list(sav4br,sav6br,sav8br,sav11br),fill = TRUE, use.names = TRUE)

```







```{r include=FALSE}
 #define factors for 1-4 
#saltset
ndns_1_11[, SaltChk := factor(SaltChk, levels = 1:8, labels = c("Salt",
	"Salt substitute",
	"Neither",
	"Item not applicable",
	"No answer/refused",
	"Don't know",
	"Qn not applicable to survey year",
	"Schedule not applicable"))]
 ndns_1_11[, SalHowC := factor(SalHowC, levels = 1:8,labels = c("Always",
"Usually",
"Sometimes",
	",Item not applicable",
	"No answer/refused",
	" Don't know",
"Qn not applicable to survey year",
"Schedule not applicable"))]
 ndns_1_11[, SltSHow := factor(SltSHow,levels = 1:8,labels = c("Always",
                                                            "Usually",
                                                            "Sometimes",
                                                            ",Item not applicable",
                                                            "No answer/refused",
                                                            " Don't know",
                                                            "Qn not applicable to survey year",
                                                            "Schedule not applicable"))]
 
 ndns_1_11[, vegetarn := factor(vegetarn, levels = 1:3, labels = c("vegetarian","vegan","not vegetarian"))]
```
 
 
```{r include=FALSE}
#medset
 ndns_1_11[, bpmedc2 := factor(bpmedc2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
 ndns_1_11[, bpmedd2 := factor(bpmedd2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, diur2 := factor(diur2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, beta2 := factor(beta2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, aceinh2 := factor(aceinh2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, calciumb2 := factor(calciumb2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, obpdrug2 := factor(obpdrug2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, PregNowB := factor(PregNowB,levels = 1:2,	labels = c("preg or breastfeeding"
	,"not preg or Bfeed"))]
 
 


 
 
 
#incset
  
 
ndns_1_11[,gor := factor(gor, levels = 1:12, labels = c("North East","North West","Yorkshire & The Humber","East Midlands","West Midlands","East of England","London","South East","South West","Wales","Scotland","Northern Ireland"))]


ndns_1_11[,region := factor(region, levels = 1:6, labels = c("England: North","England: Central/Midlands","England: South(including London)","Scotland","Wales","Northern Ireland"))] 
  
ndns_1_11[, EIMD_2007_quintile := factor(EIMD_2007_quintile, levels = 1:5)] 
ndns_1_11[,EIMD_2010_quintile := factor(EIMD_2010_quintile, levels = 1:5)] 
ndns_1_11[,EIMD_2015_quintile := factor(EIMD_2015_quintile, levels = 1:5)] 
ndns_1_11[,educfinh := factor(educfinh, levels = 1:8)] 
ndns_1_11[,nssec8 := factor(nssec8, levels = 1:8)] 
#bpset 
 
ndns_1_11[, "omsysval" := as.numeric(omsysval)]


 #hypeset
ndns_1_11[,hyper140_2 := factor(hyper140_2,levels = 1:9, labels = c(" Normotensive untreated",
" Normotensive treated",
	"Hypertensive treated",
"Hypertensive untreated",
"No answer/refused",
	"Don't know",
"Refused, attempted but not obtained, not attempted",
	 "Qn not applicable to survey year",
"Item not applicable")) ]
 ndns_1_11[, hibp140_2 := factor(hibp140_2, levels = 1:7, labels = c("Not high BP",
"High BP",
	"No answer/refused",
	"Don't know",
	"Refused, attempted but not obtained, not attempted",
	"Qn not applicable to survey year",
	"Item not applicable")) ]
ndns_1_11[, hyper1_2 := factor(hyper1_2,levels = 1:9, labels = c(" Normotensive untreated",
                                                          " Normotensive treated",
                                                          "Hypertensive treated",
                                                          "Hypertensive untreated",
                                                          "No answer/refused",
                                                          "Don't know",
                                                          "Refused, attempted but not obtained, not attempted",
                                                          "Qn not applicable to survey year",
                                                          "Item not applicable"))]
 ndns_1_11[, highbp1_2 := factor(highbp1_2, levels = 1:7, labels = c("Not high BP",
                                                              "High BP",
                                                              "No answer/refused",
                                                              "Don't know",
                                                              "Refused, attempted but not obtained, not attempted",
                                                              "Qn not applicable to survey year",
                                                              "Item not applicable")) ]
 
 
 
 
  
#ethnset  

ndns_1_11[ , ethgrp5 := factor(ethgrp5, levels = 1:5, labels = c( 'White'
                                                             , 'Mixed ethnic group'
                                                             , 'Black or Black British'
                                                             , 'Asian or asian British'
                                                             , 'Any other group'))]
ndns_1_11[ , ethgrp2 := factor(ethgrp2, levels = 1:2, labels = c( 'White'
                                                             , 'Non-white'))]




#ageset
ndns_1_11[,agegad1 := factor(agegad1, levels = 1:4 ,labels = c("16-24","25-49","50-64","65+ years"))]
ndns_1_11[,agegad2 := factor(agegad2, levels = 1:5, labels = c("16-18","19-34","35-49","50-64","65+ years"))]
ndns_1_11[,agegch1 := factor(agegch1, levels = 1:3, labels = c("8-10","11-12","13-15"))]
ndns_1_11[,agegr1 := factor(agegr1, levels = 1:5, labels = c("1.5-3 years","4-10 years","11-18 years","19-64 years","65+ years"))]
 




 
```










### Data preparation

The data is then arranged into a format which allows processing.
This includes identifying continuous and categorical variables. 
It also includes naming the categories of the categorical variables.

 
 
 
```{r include=FALSE}
#this identifies the variables wanted from the person level data sets
perslevdat <- c("seriali","Age","SurveyYear", "Sex","Country","Sodiummg","Calciummg","VitaminDµg","TotalEMJ", "Totalsugarsg","Glucoseg","Sucroseg","Fructoseg","Lactoseg","SOFTDRINKSLOWCALORIE","SOFTDRINKSNOTLOWCALORIE", "TEACOFFEEANDWATER")



```








The data is then combined into two comprehensive tables.
```{r message=FALSE, warning=FALSE, include=FALSE}
#building the tables
sav4r <- sav4r[, ..perslevdat]
sav6r <- as.data.table(sav6r[, ..perslevdat])
sav8r <- as.data.table(sav8r[, ..perslevdat])
sav11r <- sav11r[,..perslevdat]


val_labels(sav8r) <- NULL
val_labels(sav6r) <- NULL
val_labels(sav4r) <- NULL
val_labels(sav11r) <- NULL

ndns_1_11r <- rbindlist(list(sav4r,sav6r,sav8r,sav11r),fill = TRUE, use.names = TRUE)

ndns_1_11 <- ndns_1_11[ ndns_1_11r, on = "seriali"]



```





```{r echo=FALSE}

ndns_1_11[, Sex := factor(Sex, levels = 1:2, labels = c("Male", "Female"))]
ndns_1_11[, Country := factor(Country)]



```


 
The food diary data needs more processing. In particular the NOVA categorisation is not in the data set. 
I have derived UPFNOVA from a paper which had a data table identifying the NDNS sub food groups by Rauber et al.



```{r load food diary data, message=FALSE, warning=FALSE, include=FALSE}
UPFNOVA <- data.table(read_excel("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/altquant/UPFNOVA.xlsx"))
#ndns_yr1_nutrientdatabank <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr1_nutrientdatabank.sav"))

#ndns_yr11_nutrient <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr11_nutrientdatabank_2021-03-19.sav"))

 ndns1_4a_foodleveldietarydata_uk_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_foodleveldietarydata_uk_v2.sav"))

ndns5_6a_foodleveldietarydata_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_foodleveldietarydata_v2.sav"))
 
ndns7_8a_foodleveldietarydata_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_foodleveldietarydata.sav"))



 ndns_rp_yr9a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr9a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr10a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr10a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr11a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr11a_foodleveldietarydata_uk_20210831.sav") )
```

####  Processing the food diaries
The Nova group is attached to the foods in the food diaries from Rauber et al @rauberUltraprocessedFoodsExcessive2019b.
The tables are reduced to the necessary variables.

```{r process food data into tables, message=FALSE, warning=FALSE, include=FALSE}
#taking data from diet diary

UPFNOVA[,`Nova group`:= factor(`Nova group`,levels = 1:4)]
UPFNOVA$SubFoodGroupCode <- UPFNOVA$`food group number`

#specifically coded food groups

#combine with key for Nova Group
#first 1-4
foodlevedat4s <- ndns1_4a_foodleveldietarydata_uk_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
#foodlevedat4s[ ,seriali  := factor(seriali)]
foodlevedat4s<- foodlevedat4s[UPFNOVA, on= "SubFoodGroupCode"]
#summary(foodlevedat4s)

foodlevedat4s<- foodlevedat4s[,c("seriali","TotalGrams","Nova group","EnergykJ")]

#years 5-6
foodlevedat6s <- ndns5_6a_foodleveldietarydata_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat8s <- ndns7_8a_foodleveldietarydata_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat6s<- foodlevedat6s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat8s<- foodlevedat8s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat6s<- foodlevedat6s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat8s<- foodlevedat8s[,c("seriali","TotalGrams","Nova group","EnergykJ")]


#finally years 9, 10, 11
foodlevedat9s <- ndns_rp_yr9a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat10s <- ndns_rp_yr10a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat11s <- ndns_rp_yr11a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]

#combine with key for Nova Group

foodlevedat9s<- foodlevedat9s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat10s<- foodlevedat10s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat11s<- foodlevedat11s[UPFNOVA, on= "SubFoodGroupCode"]



foodlevedat9s<- foodlevedat9s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat10s<- foodlevedat10s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat11s<- foodlevedat11s[,c("seriali","TotalGrams","Nova group","EnergykJ")]

```

```{r message=FALSE, warning=FALSE, include=FALSE}
#heatmap(foodlevedat9s, Colv = "TotalGrams")

```


To work out the gram weight amount of food intake by each individual, first the diary entries for each individual are totalled up.
The total gram weight value of intake of each food is then worked out as a percentage of the total intake.

```{r function defn, message=FALSE, warning=FALSE, include=FALSE}
## First I write the function to process the individual entries. This totals the intake in grams and identifies the proportion by nova type.



IndividDiet <- function(DT,individ) {
  pers <- DT[seriali==individ]
pers
oneweight<-pers[, round(sum(TotalGrams),2), by=`Nova group`][order(`Nova group`)]

setnames( oneweight, 'V1','Totalg')
 oneweight[,pcnt := round((Totalg/sum(Totalg)*100),2)]

  return(na.exclude(oneweight))
}


```

All these individual calculations are then built back up into tables. 
This is done for years 9-11 and then 1-4.


```{r year 9 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat9s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat9s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat9sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat9sp



```

```{r year 10 process, warning=FALSE, include=FALSE}

name <-foodlevedat10s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat10s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat10sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat10sp


```


```{r year 11 process, message=FALSE, warning=FALSE, include=FALSE}

name <-foodlevedat11s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat11s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat11sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat11sp

#then combining the outputs 9-11
foodleveldat910sp <- rbind(foodleveldat9sp,foodleveldat10sp)
foodleveldat911sp <- rbind(foodleveldat11sp, foodleveldat910sp)
```


```{r year 5-6 process, warning=FALSE, include=FALSE}

name <-foodlevedat6s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat6s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat6sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat6sp


```


```{r year 7-8 process, warning=FALSE, include=FALSE}

name <-foodlevedat8s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat8s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat8sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat8sp


```


```{r slow processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat4s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat4s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat4sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat4sp



```

The process can be done for food level energy intake also.
```{r energy function defn, message=FALSE, warning=FALSE, include=FALSE}
## First I write the function to process the individual entries. This totals the intake in grams and identifies the proportion by nova type.



IndividDietE <- function(DT,individ) {
  pers <- DT[seriali==individ]
pers
oneweight<-pers[, round(sum(EnergykJ),2), by=`Nova group`][order(`Nova group`)]

setnames( oneweight, 'V1','EnergykJ')
 oneweight[,Epcnt := round((EnergykJ/sum(EnergykJ)*100),2)]

  return(na.exclude(oneweight))
}


```

```{r Eyear 9 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat9s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat9s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat9spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat9spE



```

```{r Eyear 10 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat10s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat10s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat10spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat10spE



```


```{r Eyear 11 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat11s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat11s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat11spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
foodleveldat11spE


#then combining the outputs 9-11
foodleveldat910spE <- rbind(foodleveldat9spE,foodleveldat10spE)
foodleveldat911spE <- rbind(foodleveldat11spE, foodleveldat910spE)
```


```{r Eslow processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat4s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat4s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat4spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat4spE



```


```{r E8 processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat8s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat8s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat8spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat4spE



```

```{r E6 processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat6s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat6s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat6spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat6spE



```


After that, this information is added to the other data.
This gives us the nova group information by weight and weight percent for all participants .

```{r adding to earlier data2, echo=FALSE, message=FALSE, warning=FALSE}
# combine with data from version10a.rmd

#sav11rpb <- readRDS("sav11rp.rds")


setkey(foodleveldat911sp, seriali)
setkey(foodleveldat911spE, seriali)

foodleveldat911spE <-foodleveldat911spE[foodleveldat911sp, by= seriali]
#sav11rp <-sav11rp[foodleveldat911spE, by= seriali]
#sav11rp$ED <- sav11rp[,.("Epcnt_4"/"pcnt_4")]

#sav4rp <- readRDS("sav4rp.rds")

#setkey(sav4rp,seriali)
setkey(foodleveldat4sp,seriali)
setkey(foodleveldat4spE,seriali)
foodleveldat4spE<-foodleveldat4spE[foodleveldat4sp, by= seriali]
#sav4rp <-sav4rp[foodleveldat4spE, by= seriali]


#setkey(sav6rp,seriali)
setkey(foodleveldat6sp,seriali)
setkey(foodleveldat6spE,seriali)
foodleveldat6spE <-foodleveldat6spE[foodleveldat6sp, by= seriali]
#sav6rp <-sav6rp[foodleveldat6spE, by= seriali]



#setkey(sav8rp,seriali)
setkey(foodleveldat8sp,seriali)
setkey(foodleveldat8spE,seriali)
foodleveldat8spE <-foodleveldat8spE[foodleveldat8sp, by= seriali]
#sav8rp <-sav8rp[foodleveldat8spE, by= seriali]

ndns_1_11food <- rbindlist(list(foodleveldat8spE,foodleveldat4spE,foodleveldat6spE,foodleveldat911spE))
```



The data is now ready for analysis first by descriptive analysis.





```{r adding to earlier data, echo=FALSE, message=FALSE, warning=FALSE}
# combine with data from version10a.rmd
setkey(ndns_1_11food, seriali)
setkey(ndns_1_11, seriali)
ndns_1_11 <- ndns_1_11[ndns_1_11food, by= seriali]
write(ndns_1_11,"ndns_1_11")
```