---
title: "Exposure to High Ultra-processed Food and Sodium Intake and its effect on Hypertension from a Cross-sectional study  UK National Dietary and Nutritional Survey (NDNS) in England 2008-2019"
output:
  html_notebook: 
    toc: yes
    fig_caption: yes
    number_sections: yes
    code_folding: hide
  pdf_document: 
    toc: yes
    fig_caption: yes
    number_sections: yes
    keep_tex: yes
    latex_engine: xelatex
    fig_width: 5
    fig_height: 5
  word_document: 
    toc: yes
    fig_caption: yes
    keep_md: yes
    toc_depth: 4
    fig_width: 5
    fig_height: 5
  html_document: 
    toc: yes
    fig_caption: yes
    number_sections: yes
    keep_md: yes
    fig_width: 4
    fig_height: 4
always_allow_html: true 
editor_options: 
  markdown: 
    wrap: sentence
    bibliography: upfnabp.bib 
bibliography: references.bib
csl: Vancouver.csl
---

# Exposure to High Ultra-processed Food and Sodium Intake and its effect on Hypertension from a cross sectional study UK National Dietary and Nutritional Survey (NDNS) in England 2008-2019 {.unnumbered}

# David O'Hagan {.unnumbered}

200299857

## Dissertation submitted in partial fulfilment of the requirements for the degree of Master of Public Health, The University of Liverpool {.unnumbered}

## August 2023 {.unnumbered}

\newpage

no portion of this work has been submitted in support of an application\
for degree or qualification of this or any other University or institute of learning.\
Signature

\newpage

## Dedication {.unnumbered}

To Julie Andrew and Sophie

for your loving patience and support

\newpage

# Abstract {.unnumbered}

Hypertension is associated with exposure to high intake of 'industrialy produced edible substances' or UPF, and high Sodium intake.
This research examines assumptions about salt and UPF.
What contributes to 'unhealthiness'?
Will 'reformulation' generate 'healthy UPF'?
Can this help reduce BP in one part of the the UK?

### Method

This research uses cross-sectional data with a stratified sample representative of the UK population.
It is secondary data analysis of the National Dietary and Nutrition Survey (NDNS 2008-2019).

Using multivariable logistic regression analysis of high sodium intake and high UPF intake against hypertension, secondary end points included regression of Sodium intake against UPF intake, age against BP, sodium intake and UPF intake.

### Results

There was an increased odds ratio of hypertension with higher sodium intake (OR=5.57(1.47,21.2)).

There was a lower odds ratio of hypertension with high upf intake (OR=0.57(0.34,0.94)).

There was no correlation between UPF intake and sodium intake (beta=0).

There is a strong correlation between age and BP (beta =0.43 (CI 0.41,0.45)), as well as age and Sodium intake.

There is a strong age gradient of UPF intake (beta= -0.25 (CI -0.26,-0.23)).

### Conclusion

This study shows that high Na intake is associated with hypertension.
Reduction of sodium intake may be effective at reducing the overall risk.

UPF intake is also associated with hypertension.
This may show reverse causation in this cross-sectional study.

Policy should aim to reduce intake of sodium, longitudinal studies may be more effective at identifying the causal relationship between UPF and BP.

The lack of association between UPF and sodium intake is odd.
Most UPF contains more sodium.
Some UPF contains less sodium.
This result suggests that the mix of UPF consumed by this population has no net increase in sodium content.
If this is true, reformulation for low salt would not eliminate the association identified in the main results.
Public health policy will need to reduce UPF, not simply reformulate.

### Keywords

UPF, Sodium, hypertension, reformulation, Nutrition

**Abstract** 296 words

**Dissertation** 9676

\newpage

## Acknowledgments {.unnumbered}

Thanks to Zoe and Martin

Thanks to Paul for the project which didn't quite come together

\newpage

```{r library, include=FALSE}
library(haven)
library(data.table)
library(readxl)
#library(sensemakr)
library(labelled)
library(survey)
library(srvyr)
library(gtsummary)
library(stargazer)
library(knitr)
library(flextable)
library(officer)
library(dagitty)
library(grateful)
library(dplyr)
library(DiagrammeR)
library(RefManageR)
```

```{=tex}
\tableofcontents
\listoffigures
\listoftables
\newpage
```
```{r tblabbreviations, echo=FALSE, message=TRUE, warning=TRUE}
#| tab-id: "table 1"
#| caption: "Table of abbreviations used"

table1 <- data.table("Abbreviation "=c("AIC","BMI","BP","CI","CHAMPs","CVD","HR","IMD","Na","NCD","NDNS","NOVA","OR","UPF" ),"Term"=c("Akaike Information Criterion","Body Mass Index","Blood Pressure","Confidence intervals 95% level","Cheshire and Merseyside public health  collaborative","Cardiovascular Disease", "Hazard ratio","Index of Multiple Deprivation","Sodium intake in mg",  "Non communicable Disease","National Dietary and Nutrition Survey","NOVA is a classification system, it is not an acronym","Odds Ratio","Ultra Processed Foods"))
table1 <- width(autofit(flextable(table1)), width=5)
table1 <- set_caption(table1, "Table of Abbreviations used", 
  autonum = run_autonum(seq_id = "tab", bkm = "table1"))
table1
```

\newpage

# Introduction

417482 people, 15.4% of the population, have hypertension in Cheshire and Merseyside @cardiova.
Cheshire and Merseyside public health collaborative (CHAMPs) have a plan to reduce blood pressure (BP) by 2029 @highblo2020 .
The strategy aims to increase 'awareness'.
This is intended to increase individual compliance with testing and treatment of raised BP.
This study intends to offer additional opportunities for improving outcomes.

UPF makes up up to 60% diet in UK especially in the North West of England.
There is evidence of an association between hypertension and intake of Ultra-processed Foods (UPF) @du2014 and hypertension and Salt intake @cappuccio2015 from studies of different types in multiple countries.
Moreira et al @moreira2015 model a scenario where "halving intake of (NOVA) group 3 foods could result in approximately 22,055 fewer deaths" across the UK in 2030.
Local food policies such as that of Liverpool @liverpoo2018 can make a significant public health impact.
Byker-Shanks @bykershanks2022 identify the effect it can have on UPF intake, and so might be a way of reducing hypertension at a population level.

Marmot @marmot2022 identifies the external influences in Cheshire and Merseyside which need to be improved to permit individual action to be effective.
Using a better understanding of the role of UPF, and the interaction with sodium, might give a mechanism of action of some of Marmot's categories of influence.

This study assessed prevalence of exposure to sodium and UPF, and hypertension using a nationally representative cross-sectional study, the National Dietary and Nutrition Survey (NDNS) @universityofcambridge2022 .This study has data from the UK from 2008 to 2019.
It is stratified to be representative of the population of the UK by sex, age, region and index of multiple deprivation (IMD).
BP, UPF and Sodium intake were recorded.
Age and sex remain important background factors.

I have used STROBE guidance @elm2007 in producing this report.
This study explored this complex web pulling out strands within it, [diagram 1](fig:diagram%201) shows a possible arrangement of this.

```{r figdiag1, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-id: "diagram 1"
#| fig-cap: "The relationships explored in the analysis"

bp1 <- dagitty(" dag {
               Sodium -> Blood_Pressure
               Age -> Blood_Pressure
               Sex ->  Blood_Pressure
               UltraProcessed_Food ->  Blood_Pressure
             UltraProcessed_Food  <-> Sodium
               Age ->  UltraProcessed_Food
               Sex ->  UltraProcessed_Food 
               Blood_Pressure -> non_communicable_Disease
               Age -> Sodium
               Sex -> Sodium
               
            
}")
coordinates(bp1) <- list(x=c(Sodium =1, Ultraprocessed_Food=2, Sex=3,Age=4,Blood_Pressure=5, non_communicable_Disease=6),y=c(Sodium =1, Ultraprocessed_Food=2, Sex=3,Age=4,Blood_Pressure=5, non_communicable_Disease=6))
plot(graphLayout(bp1))

```

## Epistemology

The epistemological approach of this study is positivist.
I use a quantitative approach in a mechanistic and deterministic model.
However, I am aware that this model is incomplete.
Positivism encourages experimental isolation, control of the experimental environment.

Real world dietary change requires understanding interaction with cultural,social and economic factors, not isolation.
Critical realist and social constructionist studies are needed to complement the information from this study.
The commercial and social determinants of health are models which have a great deal of impact on exposure to UPF and sodium, and on dietary effects on BP.

## Positionality

In a positivist paradigm the observer is external to the experiment.
Acknowledging the constructivist aspects to this study admits that the observer is closer to the model making my positionality of interest.
Jafar @jafar2018 argues that understanding the position of the investigator is of interest to understanding quantitative study.

I bring an attachment to positivist ideals from my biomedical background.
As an older physician I am aware of social factors impacting health of participants as Evans and Trotter @evans discuss.
I also understand that my perception of the world is from a position of significant privilege.
To proceed, I need to be aware of the limitations of the positivist approach, and of these aspects of my view of the world.

These constructed ideas, social expectations, income, or geography affect food and health 'choices'.
They also impact on 'hard' clinical measurements such as BP, through physical position and room temperature as well as by the relationship between the observer and the participant.

This work is primarily to complete requirements for an MPH degree which means that it is influenced by factors around health equity and classic epidemiology as taught on the course.
It is produced in collaboration with a research group with a long established reputation in food research in public health, which may steer the results in a conservative direction.

\newpage

# Literature Review

## Rationale

-   1 develop search
-   2 review search and confirm inclusion
-   3 describe literature
-   4 synthesise literature
-   5 critique literature
-   6 explain role of study within context

This literature review was intended as a systematic search, to identify papers with information about UPF, sodium and blood pressure informed by PRISMA @page2021.

The rationale for this review is to contribute background and to answering the research question;

What is the evidence that in adults and children across the four home nations of the UK between 2008 and 2019, would exposure to high sodium dietary intake, and or high UPF dietary intake, compared to lower exposure, increase the odds of having a mean systolic blood pressure of over 140mmHg?

## Method

Eligible studies were cross-sectional studies, and systematic reviews considering the relationship between the exposure, and outcome in comparable general populations.
Papers were excluded where the population was specifically of one type, or had a specified health condition.
Another exclusion criterion was where specific foods were considered.

Scopus @scopus2 , Pubmed , Web of Science and Medline(ovid) @ovid:se were searched.
Grey literature, especially government reports around NDNS were studied, but they did not fit the inclusion criteria for this literature review.
The Cochrane database of systematic reviews was accessed.

The search strategy is included in [table 1](#tab:table2) below.
The search concentrated on systematic reviews, these identified high quality primary research.
Other sources were identified by cross referencing bibliographies particularly from the systematic reviews.
Colleagues identified further relevant literature.

My search terms

```{r table2, echo=FALSE, message=TRUE, warning=TRUE}
#| tab-id: "tab2"
#| caption: "Description of Search "


search1 <- data.table("Name"=c("Date of Search","Number of results" ),
                      "Scopus search"=c("18/6/23","240"),
                      "Medline(Ovid)Search" =c("18/6/23","202"),"Pubmed search"=c("18/6/23","103"),"Web of Science Search" =c("18/6/23","48"),"Total"=c("18/6/23","593"))
                                                                                                     # ((sodium or salt or "sodium intake") OR ("ultraprocessed food*" OR "ultra-processed food*" OR "ultra processed food*" OR "NOVA food*")) AND ( "blood pressure" OR hypertension OR "cardiovascular disease" OR "cardiovascular risk")AND(cohort OR cross-section* OR prospective or meta-analysis or "systematic review" )                                                  
table2 <-flextable(search1)
table2 <- add_header_lines(table2, values = c("search terms ((\"ultraprocessed food*\" OR \"ultra-processed food*\" OR \"ultra processed food*\" OR \"NOVA food*\")OR (salt OR sodium OR \"sodium intake\")) AND ( \"blood pressure\" OR hypertension OR \"cardiovascular disease\" OR \"cardiovascular risk\")AND(cohort OR cross-section* OR prospective OR meta-analysis OR \"systematic review\")", "Inclusion prospective and observational studies and systematic reviews specifically on hypertension with relevant exposures","Exclusions (\"renal cell cancer\" OR \"gastric cancer\" OR \"multiple sclerosis\" OR dialysis OR DHA OR auto-immunity OR autoimmunity OR diarrhoea OR telomere OR crp OR \"c-reactive\" OR CKD OR \"chronic kidney disease\" OR autosomal OR geneti* OR \"Inflammatory bowel disease\" OR diabetes OR atherosclerosis OR osteoporosis OR angiotensin*  OR aldosteron* OR covid-19 OR gestation* OR stroke OR \"birth weight\" OR hypertensive OR \"immune mechanisms\" OR \"heart failure\"OR taste OR \"cognitive decline\" OR dementia OR mortality OR validation)"))

table2 <- set_caption(table2, "Description of Search ", 
  autonum = run_autonum(seq_id = "tab", bkm = "tab2"))

table2
#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/search1.png")
#save_as_image(width(autofit(flextable(search1)), width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/search1b.png")
```

## Selection strategy

Inclusions and exclusions are identified in the table.
No time limits, language limits or availability limits were included in the search.

The papers were reviewed by the author only.

The data sought were odds ratios for the effect of UPF or sodium on blood pressure and outcomes of systematic reviews.

There is risk of bias due to the single reviewer approach.

There is some heterogeneity of approach to reporting exposure or outcome making it difficult to compare items directly.

## Search results

The search identified 593 papers, 348 after removing duplicates.
[Chart 1](fig:diag4) shows how these were assessed.

```{r diag4, echo=FALSE, message=FALSE, warning=FALSE}
searchcht <- grViz(diagram = "digraph flowchart {
  node [fontname = arial, shape = rectangle]
  tab1 [label = '@@1']
  tab2 [label = '@@2']
  tab3 [label = '@@3']
  tab4 [label = '@@4']
  tab5 [label = '@@5']
  tab6 [label = '@@6']
  
  tab1 -> tab2 -> tab3;
  tab3 -> tab4;
  tab3 -> tab5;
  tab3 -> tab6;
 
}
  
  [1]: 'Initial searches 593 (240,202,103 and 48)  papers'
  [2]: 'adjusting for overlap 348'    
  [3]: 'comparing with exclusion criteria 56'  
  [4]: 'Systematic Review and UPF 2' 
  [5]: 'Systematic Review and sodium 4' 
  [6]: 'Cohort Studies 11'
  ")
print(searchcht)
```

These were then reviewed by comparing the titles against the inclusion and exclusion criteria leaving 56.
6 systematic reviews were identified, one which considered both UPF and sodium intake.
Some of these calculated odds ratios using meta-analysis.

## Discussion of literature

+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| First Author                                | Type of Study                       | Subject of Study                | Results                                                                                                          |
+=============================================+=====================================+=================================+==================================================================================================================+
| Barbosa @barbosa2023                        | systematic review                   | UPF and Sodium and Hypertension | 5 cross-sectional and 4 cohort studies                                                                           |
|                                             |                                     |                                 |                                                                                                                  |
|                                             | Papers from Americas                |                                 | +ve association found                                                                                            |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| Wang @wangAssociationUltraprocessedFood2022 | systematic review and meta analysis | UPF and hypertension            | 5 cross-sectional and 4 cohort studies                                                                           |
|                                             |                                     |                                 |                                                                                                                  |
|                                             |                                     |                                 | (odds ratio: 1.23; 95% CI: 1.11, 1.37; P=0.034)                                                                  |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| Mambrini @mambrini2023                      | systematic review                   | UPF and BP                      | 4/17 UPF and BP                                                                                                  |
|                                             |                                     |                                 |                                                                                                                  |
|                                             | few studies one women only study    |                                 | +ve association found                                                                                            |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| D'Elia @delia2018                           | meta-analysis                       | Salt and CVD                    | 11 studies                                                                                                       |
|                                             |                                     |                                 |                                                                                                                  |
|                                             |                                     |                                 | *a 2.84% (95% CI 0.51--5.08) reduction in PWV*                                                                   |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| Leyvraz @leyvraz2018                        | meta-analysis                       | sodium blood pressure           | 13/18 studies                                                                                                    |
|                                             |                                     |                                 |                                                                                                                  |
|                                             | children                            |                                 | every additional gram of sodium intake per day, systolic blood pressure increased by 0.8 mmHg (95% CI: 0.4, 1.3) |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
| Frias @frías2023                            | systematic review of cohorts        | sodium and BP                   | 2 studies BP                                                                                                     |
|                                             |                                     |                                 |                                                                                                                  |
|                                             | children                            |                                 | no association systolic BP                                                                                       |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+
|                                             | nb BP is Blood Pressure             | nb CI is confidence interval    | nb CVD is cardiovascular disease                                                                                 |
|                                             |                                     |                                 |                                                                                                                  |
|                                             |                                     |                                 | UPF is Ultraprocessed food                                                                                       |
+---------------------------------------------+-------------------------------------+---------------------------------+------------------------------------------------------------------------------------------------------------------+

: Systematic reviews and Meta-analyses

### UPF and BP

Three systematic reviews, those of Mambrini et al,Barbosa et al, and Wang et al highlight the risk of hypertension with high UPF.
They also identify that the effect is not always evident.
Wang's meta-analysis was used to calculate the sample size for this study.
All these reviews appeared of good quality following guidance.

Mambrini et al, in their systematic review, identify that few papers attempt to link UPF with hypertension.
They identify a positive correlation.
Their systematic review identifies Monge, Scaranni, and Mendonça, whose three cohort studies of middle aged adults were followed up for between 2.2 and 9.1 years.

Scaranni used Brazil's ELSA study, in middle aged civil servants, finding that higher UPF had a marginally greater risk of developing hypertension (OR = 1.17; 95% CI: 1.00, 1.37) compared with lower intake.

Monge found no association between categories of UPF, from ≤20% to \>45% energy/d and incident hypertension.
This study used mexican female teachers, this group may have a different exposure and outcome profile to a more general population.

Mendonça in Spain found an affect on hypertension (HR = 1.21, 95% CI: 1.06, 1.37, Ptrend = 0.004) in adults.
This sample had a strong effect on Wang's meta analysis with a weighting of 16.72 from 9 studies.

Wang's meta analysis included six more studies; Ivancovsky-Wajcman OR = 1.53, (1.07- 2.19) in the USA, Levigne-Robichaud OR =0·99 (0·59, 1·68) in Canada, Martinez-Steele OR =1.19 (1.03,1.38) in the USA, Nardocci OR = 1.60, (1.26--2.03) in Canada, Nasreddine OR =3.10, (0.84,16.66)in Lebanon, Rezende-Alves OR =1.35 (1.01,1.81) in Brazil.

Of these studies, the study by Nasreddine was a very small localised study, as shown by the wide confidence interval.
Monge and Levigne-Robichaud also had equivocal results.
Levigne-Robichaud's sample was a specific population in Canada.

Barbosa identified a different group of studies.
Conceiacao had a population of 64, Martinez-Peres 5636, Martinez-Steele 6385,and Smiljanec 40.
No meta-analysis was done.
The studies included a wider range of outcomes making it hard to compare results.
The narrative outcome was a similar positive correlation, reportedly for salt and UPF.

Shim demonstrated the effect on hypertension in Korea, despite the highest tertile percentage UPF being only \>28.55%.

+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| First Author                                         | Study Type                         | Subject                               | Results                                                                                     |
+======================================================+====================================+=======================================+=============================================================================================+
| Shim @ShimSunYoung2022Couf                           | Korean Cross-sectional             | UPF and Hypertension                  | OR= 1.25,CI: 1.11 and 1.40                                                                  |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Scaranni @ScaranniPatriciadeOliveiradaSilva2021Ufci  | Brazil ELSA cohort                 | UPF and Hypertension                  | OR = 1.17; 95% CI: 1.00, 1.37                                                               |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Rezendez-Alves @rezende-alves2021                    | Brazil cohort                      | UPF and Hypertension                  | RR: 1.35; 95% CI: 1.01, 1.82).                                                              |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Martinez-Peres @martinez-perez2021                   | Spain Transverse                   | UPF and BP                            | no significant effect                                                                       |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Monge @monge2021                                     | Mexico women only Cohort           | UPF and BP                            | no effect                                                                                   |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| da Conceicao @DaConceição2019512                     | Cross-sectional                    | UPF and BP                            | no effect                                                                                   |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Nardocci @nardocci2021                               | Canada Cross-sectional             | UPF and hypertension                  | OR = 1.60, 95% CI: 1.26--2.03)                                                              |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Nasreddine @NasreddineLara2018Ampd                   | Lebanese Cross-sectional           | UPF and BP                            | OR 3.10 (0.58,16.66)                                                                        |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Lavigne-Robichaud @Lavigne-RobichaudMathilde2018Dqii | Canada Cree People Cross-sectional | UPF and BP                            | 0.99 (0.59,1.68)                                                                            |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Martinez-Steele @martinezsteele2022                  | USA Cross-sectional                | UPF and Mets                          | 1.19(1.03,1.38)                                                                             |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Smiljanec @smiljanec2020                             | USA Cross-sectional                | UPF and BP                            | Positive association between UPFs and general SBP (B = 0.25, 95% CI: 0.03, 0.46, p = 0.029) |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| Ivancovsky-Wajcman @IvancovskyWajcmanDana2021Ufia    | USA Cross-sectional                | UPF and BP                            | OR = 1.53, 1.07- 2.19, P = .026                                                             |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| de Deus Mendonça @DeDeusMendonçaRaquel2017Ufca       | Spain cohort                       | UPF and (self -reported) hypertension | HR, 1.21; 95% CI, 1.06, 1.37; P for trend = 0.004                                           |
|                                                      |                                    |                                       |                                                                                             |
|                                                      | middle aged uni grads              |                                       |                                                                                             |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+
| nb HR is hazard ratio                                | nb BP is Blood Pressure            | nb CI is confidence interval          | nb CVD is cardiovascular disease                                                            |
|                                                      |                                    |                                       |                                                                                             |
| RR is relative risk                                  | OR is Odds ratio                   |                                       | UPF is Ultraprocessed food                                                                  |
+------------------------------------------------------+------------------------------------+---------------------------------------+---------------------------------------------------------------------------------------------+

: Table of papers on UPF and BP

### Salt and CVD

+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| First Author               | Study Type                    | Subject                      | Results                                                                                                                                                               |
+============================+===============================+==============================+=======================================================================================================================================================================+
| He @hefjm                  | meta-analysis                 | Salt and hypertension        | 11 trials in 'normotensive' 17 in 'hypertensive'                                                                                                                      |
|                            |                               |                              |                                                                                                                                                                       |
|                            | Adults some with hypertension |                              | A reduction of 100 mmol/day (6 g of salt) in salt intake predicted a fall in systolic blood pressure of ... 3.57 mmHg in normotensive individuals (systolic: Po0.001) |
+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Strazzullo @strazzullo2009 | meta-analysis                 | Salt and CVD                 | 13 studies risk of stroke                                                                                                                                             |
|                            |                               |                              |                                                                                                                                                                       |
|                            |                               |                              | (pooled relative risk higher salt intake 1.23, 95% confidence interval 1.06 to 1.43; P=0.007) and cardiovascular disease (1.14, 0.99 to 1.32; P=0.07)                 |
+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Graudal @graudal2014       | meta-analysis                 | Salt and CVD                 | 23 prospective cohort studies                                                                                                                                         |
|                            |                               |                              |                                                                                                                                                                       |
|                            | Adults                        |                              | *ACM: HR = 1.16, 95% CI = 1.03-1.30; CVDEs: HR = 1.12, 95% CI = 1.02-1.24*                                                                                            |
+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Ma @ma2022                 | meta-analysis                 | urinary sodium and CVD       | 6 prospective studies                                                                                                                                                 |
|                            |                               |                              |                                                                                                                                                                       |
|                            | Healthy adults                |                              | Each daily increment of 1000 mg in sodium excretion was associated with an 18% increase in cardiovascular risk (hazard ratio, 1.18; 95% CI, 1.08 to 1.29),            |
+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                            | nb BP is Blood Pressure       | nb CI is confidence interval | nb CVD is cardiovascular disease                                                                                                                                      |
|                            |                               |                              |                                                                                                                                                                       |
|                            |                               |                              | nb UPF is Ultraprocessed food                                                                                                                                         |
+----------------------------+-------------------------------+------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: Review papers Sodium and BP

The systematic reviews identified in the search include Barbosa, and D'Elia.
Leyvraz and Frias concentrate on effects in children where the effects were equivocal.
These reviews appeared of good quality with appropriate inclusion and exclusion criteria.

He, Graudal and Strazzullo were not identified in this search but were referenced in the other papers.

He in a meta-analysis of 11 papers, identifies reduction in BP with reduction in sodium intake.
Graudal reports that sodium reduction can go too far, identifying a 'j'-shaped curve.
He identifies the inclusion of papers with big effect sizes, and short follow up in Graudal as contributing to their identification of a negative effect of very low sodium intakes.

\
Graudal et al.
Studied cohort studies as there were no RCTs of increased sodium intake.
They found data from 23 cohort studies (n=274,683).
They showed acute cardiac events (ACM) and cerebrovascular events CVDE were increased in high sodium intake compared with usual sodium intake *(ACM: HR = 1.16, 95% CI = 1.03-1.30; CVDEs: HR = 1.12, 95% CI = 1.02-1.24)*.

Their findings identify that there might be 'too much' salt reduction possible.
They provide an explanation as to how low sodium levels may causes issues.

D'Elia et al, in their systematic review, look at arterial stiffness pressure wave velocity 'PWV' and show that this increases with salt intake.
This arterial stiffness is potentially more sensitive to sodium intake than BP.
They included 11 studies, of 14 cohorts and 431 participants studied over 1-6 weeks.
Reducing sodium intake by 89.3mmol/day was associated with 2.84% (CI0.51-5.08) reduction in PWV.

D'Elia's results show that BP is less accurately predicted than arterial stiffness.
This may be a cause of the equivocal results found by studies looking at BP.

Straluzzo et al identified 19 cohort samples, with 117025 participants followed for 3.5-19 years, resulting in over 11000 vascular events.
Identifying small, but statistically significant relative risk values.

Ma et al studied 24 hour sodium excretion.
This is more reliable than reported intake, or intake calculated from food values.
Their older group was studied for 8.8 years.
An increase of 1000 mg/day in sodium excretion was associated with an 18% increase in cardiovascular risk (hazard ratio, 1.18; 95% CI, 1.08 to 1.29).

These papers agree that sodium intake is associated with increased BP and cardiovascular issues.

## Literature review Conclusion

Of the identified papers the majority describe 2 elements out of the 3 , UPF or salt intake or BP/ CVD, this is the gap in the literature which the current is intended to address.

They show an odds ratio for CVD with raised sodium intake and hypertension with high UPF intake.
They suggests BP is an uncertain outcome measurement.
Papers often look to CVD outcomes as stronger endpoints.

Where UPF exposure has been studied with hypertension as an endpoint the cross-sectional studies have identified a link in adults, less in women or children.
The cohort studies are consistent in showing a small but measurable positive effect.
This is also identified in the meta-analyses.

This study aims to identify these two effects at the same time within a large representative cross-sectional population thereby answering this gap.
The relationship between these two effects can be shown by studying both in the same population.
This study also gives the opportunity to consider if there are associated factors and to understand the public health implications.

## Research Question

Using PICO @bruce2018 approach,

In adults and children across the four home nations of the UK between 2008 and 2019, how much exposure to high sodium dietary intake, and or high UPF dietary intake, compared to lower exposure, increased the odds of having a mean systolic blood pressure of over 140mmHg?

In addition it may be possible to consider, if there is evidence of interaction between these and is UPF or sodium most important in these changes?

## Objectives

1 Literature Review of UPF and BP, and Sodium and BP

2 Descriptive analysis of participants from NDNS with amalgamation of data across the rolling programme.

3 Analysis of exposure to UPF and sodium, and prevalence of BP \>140mmHg using regression models with associated data analysis.

4 Discussion of implications of results in answer to the research question, Public health implications and actions.
Also in relation to limitations of cross- sectional studies, and available data, as well as suggestions for further research

\newpage

# Method

## Study Setting and Design

This analysis intends to analyse the association between sodium intake, UPF intake and BP.
This is a secondary data analysis of national cross sectional data from the National Dietary and Nutritional Survey (NDNS @universityofcambridge2022).

The NDNS was commissioned in collaboration between government departments responsible for health and for food production.
Academic partners delivered reports on diet and nutrition across the United Kingdom.
The study is designed to be representative across the four home nations, and across age with balanced representation for children.
NDNS data are available via the UK national Data service for research purposes.

NDNS is a rolling cross-sectional study, in each year a new cross section of participants is enrolled from the wider population.
Questionnaires, food diaries, and nurse assessments are used to gather data.
It has been running since 2008.
The most recent data is available from 2019.
This is a rich resource with data on exposures and outcomes and explanatory variables.

## University Research Governance and Ethical Review

The ethics process for the University of Liverpool was followed and confirmation of compliance is attached at [Appendix 2 Ethics Certificate]

The storage of the data is in keeping with the research governance agreements of the University and the Data set owners.

## Participants, Inclusion and Exclusion

Participants were identified by random selection across postal units.
The sample is stratified to ensure a representative sample across the four nations (England, Wales, Scotland, and Northern Ireland) and across regions in England (North, Central/Midlands, South(including London)).
The sample is also stratified for age and sex and Index of multiple deprivation (IMD).
Participants gave informed consent, the study was conducted within ethics committee guidance.

For NDNS the intended sample is 1000 per year with 50% adults.
Each year the sample is slightly different due to differential uptake.
Oversampling is used to control this.

The relationship between salt and systolic blood pressure may be different in individuals with pathologically high BP.
Those taking BP controlling medications may have a different relationship to sodium and UPF and so were excluded for analysis.

## Exposure Variables

The participants recorded their food intake as a food diary.
Four days of the diary including a weekend day are used for each participant.
They record food and portion size as well as where food was eaten.

Based on the food and drink intake reported and with a composition data table, the NDNS team have estimated the daily intake of food by group, and large range of nutrients including sodium.

### Sodium estimation

This analysis used the daily sodium intake in mg from NDNS.
This value reflects the expected content of standard foods calculated from the diary entry.

Serum sodium values are available for the early dataset, but not the later one.
24 urinary sodium is a better indicator of dietary sodium but values are not available across the whole time period.

WHO recommended sodium intake is less than 3000mg sodium and Du et al @du2014 used categories of moderate and high intake at 5000mg and 6000mg, these categories were used for sodium.

### UPF

The NOVA classification, developed by Monteiro et al. @monteiro2010, was used to estimate the intake of UPF.
There is no record of NOVA classification in NDNS.
This was developed by comparing every food level entry in NDNS against NOVA.
A standard methodology describing the approach used has been published by Martinez-Steele et al. @martinez-steele2023 .

The energy content of the day's food was calculated by NOVA group.
This was added to the intake for the other 3 days and the total intake by NOVA group established.
The percentage of the total intake of energy was then calculated for each of the 4 Nova categories.
Nova group 4 or UPF intake (UPF) is used for this study, this dataset was provided by Dr Colombet (personal communication).

Categories of UPF intake used in other papers eg Wang @WangMei2022UFCI are low for the UK.
Centre-weighted categories were used, the central category is the mean with one standard deviation above and below.
This effectively identifies 67% in the centre of the distribution.

## Outcome Variable

BP is a quality assured mean systolic BP which is reliable across the dataset.
It was measured in mmHg using a calibrated automatic sphygmomanometer by a study nurse under specified conditions.
These conditions controlled for the effects of exercise, temperature and ill health.
The data on all these is in the dataset.
Raw BP values are also present in the dataset to allow quality review.

Hypertension is BP over 140 mmHg, categorised to enable logistic regression.
This value is identified by Du et al (@du2014) and others, though some use lower values such as 120mmHg ( @shimConsumptionUltraProcessedFood2022 )

## Other Variables

Additional explanatory variables are ones which can also influence BP.
They include Age, Sex, and BMI.
Age at completion of education , and IMD are also used.
These may have effects such as confounding, mediation, and obstruction within the analysis.

Stratification is used in the design and sampling stage to modify for these.
Including them in multivariable regression attempts to prevent them confounding the analysis.

Age is used as continuous and categorised data for the analysis.
It has a complex role in cross-sectional analysis.
It is complicated further by the stratification which ensures a large enough sample of 50/50 children and adults.

The sex of participants is identified, as an important explanatory variable.

BMI is identified as a potential explanatory variable.
There is also a known association with BP.

IMD is included to identify socio-economic patterns in the data.
This UK- nation based data is used consistently in UK studies, but has no analogue internationally.
Instead age at completion of education, or income are often used.

## Study Size

A sample size calculation for this secondary analysis is available in [Appendix 1 Approved Proposal] the initial proposal from OpenEpi @openepi2021 .
This calculated the sample size of 3526, with a ratio of 0.75 unexposed to exposed.
An intended power of 80%, at a level of statistical significance of 95% was used.
An odds ratio of 1.2 was used based on a meta-analysis by Wang et al @wangUltraProcessedFoodsConsumption2022 .

The population size in NDNS is much larger than this.
Sample size relies on having large enough subgroups exposed, and unexposed, as well as the level of the outcome.
This is why the overall sample size needs to be so large, why a national study is ideal.

## Statistical Methods

Four data batches of data ( 2008-2012, 2013-2014, 2015-2016, 2017-2019) were combined.
The data was read using 'r-studio' with the processing being carried out using packages (see appendix 3) available from CRAN @base.
In particular the package 'survey' @survey2004 was used to manage weighted data.
Generated weighting values account for differences uptake and drop out across the annual cohorts.
'Survey' also accounts for sample stratification.

### Analysis Plan

Descriptive data was tabulated to enumerate the outline structure.

The sensitivity of the data to changes in the annual cohorts was assessed.

Then the data was analysed for correlation by regression.
Aikake index statistics(AIC) were used to assess 'goodness of fit'.
(R squared does not work for logistic regression)

In all analysis P.values and confidence intervals were calculated and a value of p = 0.05 was taken as the threshold of statistical significance.

Tables of results were produced to best demonstrate the data.
For the main results univariable regression and a set of multivariable logistic regression models was developed.
Each exposure variable was modelled separately, the final model included both of the exposure variables.Multivariable regression models were constructed to manage explanatory variables which might have confounding effects on the outcome of the analysis.

```{r readdata, message=FALSE, warning=FALSE, include=FALSE}

sav4 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_personleveldietarydata_uk_v2.sav"

sav4b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_indiv_uk.sav"

sav6 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_personleveldietarydata_v2.sav"

sav6b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_indiv.sav"

sav8 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_personleveldietarydata.sav"

sav8b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_indiv.sav"



sav11 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_personleveldietarydata_uk_20210831.sav"

sav11b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_indiv_20211020.sav"

sav4r <- as.data.table(read_sav(sav4))
sav4br <-as.data.table(read_sav(sav4b))

sav6r <- as.data.table(read_sav(sav6))
sav6br <-as.data.table(read_sav(sav6b))

sav8r <- as.data.table(read_sav(sav8))
sav8br <-as.data.table(read_sav(sav8b))


sav11r <- as.data.table(read_sav(sav11))
sav11br <- as.data.table(read_sav(sav11b))
sav11br$area <- sav11br$Area
```

```{r stdnames4, message=FALSE, warning=FALSE, include=FALSE}
#standardise the variable names for grp 1-4 
sav4br$bpmedc2 <- sav4br$bpmedc
sav4br$bpmedd2 <- sav4br$bpmedd
sav4br$diur2 <- sav4br$diur
sav4br$beta2 <- sav4br$beta
sav4br$aceinh2 <- sav4br$aceinh
sav4br$calciumb2 <- sav4br$calciumb
sav4br$obpdrug2 <- sav4br$obpdrug

sav4br$ethgrp5 <- sav4br$ethgr5
sav4br$ethgrp2 <- sav4br$ethgr2

sav4br$hyper140_2 <- sav4br$hyper140
sav4br$hibp140_2 <- sav4br$hibp140
sav4br$hyper1_2 <- sav4br$hyper1
sav4br$highbp1_2 <- sav4br$highbp1

sav4br$educfinh <- sav4br$EducFin
sav6r$SurveyYear <- sav6r$Surveyyear
```

```{r stdnames11, message=FALSE, warning=FALSE, include=FALSE}
sav11br$htval <- sav11br$htval2
sav11br$wtval <- sav11br$wtval2
sav11br$bmival <- sav11br$bmival2

sav11br$gor <- sav11br$GOR
sav11br$age <- sav11br$AgeR
sav11r$Age <- sav11r$AgeR
#sav11br[omsysval <= 10, omsysval := NA]
```

```{r novafile, message=FALSE, warning=FALSE, include=FALSE }
NOVA_ForDavid <- as.data.table(read_excel("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/NOVA_ForDavid.xlsx"))

NOVA_ForDavid[, Sex := factor(Sex, levels = 1:2)]
NOVA_ForDavid[, Country := factor(Country)]
NOVA_ForDavid[, SurveyYear := factor(SurveyYear)]
NOVA_ForDavid[, Age := as.numeric(Age)]
```

```{r weight, echo=FALSE}
#Doing the weighting correction, as explain in the PDF "6533_ndns_yrs9-11_weights_guide"

n1 = sum(sav4br$wti_UKY1234)  #Should be 6828

n2 = sum(sav6br$wti_Y56)         #Should be 2546

n3 = sum(sav8br$wti_Y78)         #Should be 2723

n4 = sum(sav11br$wti_Y911)       #Should be 3558

N <- sum(n1, n2, n3, n4) #Should be 15655
```

```{r weight2,message=FALSE, warning=FALSE, include=FALSE}
#now the proportion weight for each is calculated
sav4br$wti_UKY1to11 <- ((sav4br$wti_UKY1234/n1) * N)  * (4/11)

sav6br$wti_UKY1to11 <- ((sav6br$wti_Y56/n2) * N) * (2/11)

sav8br$wti_UKY1to11 <- ((sav8br$wti_Y78/n3) * N) * (2/11)

sav11br$wti_UKY1to11 <- ((sav11br$wti_Y911/n4) * N) * (3/11)

```

```{r allsetupvariables, message=FALSE, warning=FALSE, include=FALSE}
 #measured and recorded data yr1-4a_indiv_uk 
#subsets of the table to identify grouped information
bpset <- c("seriali","omsysval","omdiaval")
ethnset <- c("ethgrp5","ethgrp2")
saltset <- c("SaltChk","SalHowC","SltSHow" ,"vegetarn")
medsset <- c( "bpmedc2","bpmedd2","diur2","beta2","aceinh2", "calciumb2","obpdrug2", "PregNowB")
hypset <- c("hyper140_2", "hibp140_2", "hyper1_2", "highbp1_2")
incset <-c("nssec8","gor","region", "EIMD_2007_quintile","EIMD_2010_quintile","EIMD_2015_quintile","WIMD_2014_quintile","SIMD_2016_quintile","NIMD_2017_quintile","educfinh")
measset <- c("htval","wtval","bmival")
ageset <- c("agegad1", "agegad2", "agegch1","agegr1","age")
weightset <- c("wti_UKY1to11","astrata1","area")

indiv_data <- c(bpset,ethnset,saltset,medsset,hypset,incset,measset,ageset,weightset)
```

```{r individdata, message=FALSE, warning=FALSE, include=FALSE}
sav11br <- sav11br[,..indiv_data]
sav4br <- sav4br[,..indiv_data]
sav6br <- sav6br[,..indiv_data]
sav8br <- sav8br[,..indiv_data]

val_labels(sav8br) <- NULL
val_labels(sav6br) <- NULL
val_labels(sav4br) <- NULL
val_labels(sav11br) <- NULL
#sav8br$seriali <- as.double(sav8br$seriali)
#sav8br$ethgrp5 <- as.double(sav8br$ethgrp5)
```

```{r joiningtogether,message=FALSE, warning=FALSE, include=FALSE}
#this creates the whole table
ndns_1_11 <- rbindlist(list(sav4br,sav6br,sav8br,sav11br),fill = TRUE, use.names = TRUE)

```

```{r definefactors,include=FALSE}
 #define factors for 1-4 
#saltset
ndns_1_11[, SaltChk := factor(SaltChk, levels = 1:8, labels = c("Salt",
	"Salt substitute",
	"Neither",
	"Item not applicable",
	"No answer/refused",
	"Don't know",
	"Qn not applicable to survey year",
	"Schedule not applicable"))]
 ndns_1_11[, SalHowC := factor(SalHowC, levels = 1:8,labels = c("Always",
"Usually",
"Sometimes",
	",Item not applicable",
	"No answer/refused",
	" Don't know",
"Qn not applicable to survey year",
"Schedule not applicable"))]
 ndns_1_11[, SltSHow := factor(SltSHow,levels = 1:8,labels = c("Always",
                                                            "Usually",
                                                            "Sometimes",
                                                            ",Item not applicable",
                                                            "No answer/refused",
                                                            " Don't know",
                                                            "Qn not applicable to survey year",
                                                            "Schedule not applicable"))]
 
 ndns_1_11[, vegetarn := factor(vegetarn, levels = 1:3, labels = c("vegetarian","vegan","not vegetarian"))]
```

```{r definefactors2 ,include=FALSE}
#medset
 ndns_1_11[, bpmedc2 := factor(bpmedc2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
 ndns_1_11[, bpmedd2 := factor(bpmedd2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, diur2 := factor(diur2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, beta2 := factor(beta2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, aceinh2 := factor(aceinh2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, calciumb2 := factor(calciumb2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, obpdrug2 := factor(obpdrug2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, PregNowB := factor(PregNowB,levels = 1:2,	labels = c("preg or breastfeeding"
	,"not preg or Bfeed"))]
 
 


 
 
 
#incset
  
 
ndns_1_11[,gor := factor(gor, levels = 1:12, labels = c("England:North East","England:North West","England:Yorkshire & The Humber","England:East Midlands","England:West Midlands","England:East of England","England:London","England:South East","England:South West","Wales","Scotland","Northern Ireland"))]


ndns_1_11[,region := factor(region, levels = 1:6, labels = c("England: North","England: Central/Midlands","England: South(including London)","Scotland","Wales","Northern Ireland"))] 
  
ndns_1_11[, EIMD_2007_quintile := factor(EIMD_2007_quintile, levels = 1:5)] 
ndns_1_11[,EIMD_2010_quintile := factor(EIMD_2010_quintile, levels = 1:5,labels = c(Label = "0.53->8.49 [least deprived]", "8.49->13.79","13.79->21.35","21.35->34.17", "34.17->87.80 [most deprived]"))] 
ndns_1_11[,EIMD_2015_quintile := factor(EIMD_2015_quintile, levels = 1:5,labels = c(Label = "Most deprived", "2","3","4", "least deprived"))] 
ndns_1_11[,SIMD_2016_quintile := factor(SIMD_2016_quintile, levels = 1:5,labels = c(Label = "Most deprived", "2","3","4", "least deprived"))]
ndns_1_11[,WIMD_2014_quintile := factor(WIMD_2014_quintile, levels = 1:5,labels = c(Label = "Most deprived", "2","3","4", "least deprived"))]
ndns_1_11[,NIMD_2017_quintile := factor(NIMD_2017_quintile, levels = 1:5,labels = c(Label = "Most deprived", "2","3","4", "least deprived"))]

ndns_1_11[,educfinh := factor(educfinh, levels = 1:8, labels = c( "Not yet finished",
"Never went to school"
,"14 or under"
,"15","16","17","18","19 or over"))] 
ndns_1_11[,nssec8 := factor(nssec8, levels = 1:8)] 
#bpset 
 
#ndns_1_11[, "omsysval" := as.numeric(omsysval)]


 #hypeset
ndns_1_11[,hyper140_2 := factor(hyper140_2,levels = 1:9, labels = c(" Normotensive untreated",
" Normotensive treated",
	"Hypertensive treated",
"Hypertensive untreated",
"No answer/refused",
	"Don't know",
"Refused, attempted but not obtained, not attempted",
	 "Qn not applicable to survey year",
"Item not applicable")) ]
 ndns_1_11[, hibp140_2 := factor(hibp140_2, levels = 1:7, labels = c("Not high BP",
"High BP",
	"No answer/refused",
	"Don't know",
	"Refused, attempted but not obtained, not attempted",
	"Qn not applicable to survey year",
	"Item not applicable")) ]
ndns_1_11[, hyper1_2 := factor(hyper1_2,levels = 1:9, labels = c(" Normotensive untreated",
                                                          " Normotensive treated",
                                                          "Hypertensive treated",
                                                          "Hypertensive untreated",
                                                          "No answer/refused",
                                                          "Don't know",
                                                          "Refused, attempted but not obtained, not attempted",
                                                          "Qn not applicable to survey year",
                                                          "Item not applicable"))]
 ndns_1_11[, highbp1_2 := factor(highbp1_2, levels = 1:7, labels = c("Not high BP",
                                                              "High BP",
                                                              "No answer/refused",
                                                              "Don't know",
                                                              "Refused, attempted but not obtained, not attempted",
                                                              "Qn not applicable to survey year",
                                                              "Item not applicable")) ]
 
 
 
 
  
#ethnset  

ndns_1_11[ , ethgrp5 := factor(ethgrp5, levels = 1:5, labels = c( 'White'
                                                             , 'Mixed ethnic group'
                                                             , 'Black or Black British'
                                                             , 'Asian or asian British'
                                                             , 'Any other group'))]
ndns_1_11[ , ethgrp2 := factor(ethgrp2, levels = 1:2, labels = c( 'White'
                                                             , 'Non-white'))]




#ageset
ndns_1_11[,agegad1 := factor(agegad1, levels = 1:4 ,labels = c("16-24","25-49","50-64","65+ years"))]
ndns_1_11[,agegad2 := factor(agegad2, levels = 1:5, labels = c("16-18","19-34","35-49","50-64","65+ years"))]
ndns_1_11[,agegch1 := factor(agegch1, levels = 1:3, labels = c("8-10","11-12","13-15"))]
ndns_1_11[,agegr1 := factor(agegr1, levels = 1:5, labels = c("1.5-3 years","4-10 years","11-18 years","19-64 years","65+ years"))]
 




 
```

```{r personaldatadefn,include=FALSE}
#this identifies the variables wanted from the person level data sets
perslevdat <- c("seriali","Age","SurveyYear", "Sex","Country","Sodiummg","Calciummg","VitaminDµg","TotalEMJ", "Totalsugarsg","Glucoseg","Sucroseg","Fructoseg","Lactoseg","SOFTDRINKSLOWCALORIE","SOFTDRINKSNOTLOWCALORIE", "TEACOFFEEANDWATER")



```

```{r joinpersleveldata,message=FALSE, warning=FALSE, include=FALSE}

val_labels(sav8r) <- NULL
val_labels(sav6r) <- NULL
val_labels(sav4r) <- NULL
val_labels(sav11r) <- NULL

 sav4r$Sex <- as.numeric(sav4r$Sex)
sav6rSex <- as.numeric(sav6r$Sex)

#building the tables
sav4r <- sav4r[, ..perslevdat]
sav6r <- as.data.table(sav6r[, ..perslevdat])
sav8r <- as.data.table(sav8r[, ..perslevdat])
sav11r <- sav11r[,..perslevdat]


val_labels(sav8r) <- NULL
val_labels(sav6r) <- NULL
val_labels(sav4r) <- NULL
val_labels(sav11r) <- NULL

ndns_1_11r <- rbindlist(list(sav4r,sav6r,sav8r,sav11r),fill = TRUE, use.names = TRUE)

ndns_1_11 <- ndns_1_11[ ndns_1_11r, on = "seriali"]



```

```{r tidyup, echo=FALSE}

ndns_1_11[, Sex := factor(Sex, levels = 1:2, labels = c("Male", "Female"))]
ndns_1_11[, Country := factor(Country)]
ndns_1_11[, SurveyYear := factor(SurveyYear)]
ndns_1_11[omsysval <= 10, omsysval := NA]


upfd <- c("seriali","NOVA_Epct_opti_4","NOVA_Gpct_opti_4")
NOVA_ForDavid <- NOVA_ForDavid[,..upfd]
#NOVA_ForDavid
ndns_1_11 <- ndns_1_11[NOVA_ForDavid, on = "seriali"]
```

```{r newvariables, echo=FALSE,message=FALSE}
#epcnt_4 divided into tertiles as cat variable
#ndns_1_11$UPF <-  ndns_1_11$Epcnt_4 
                      
#ndns_1_11$UPF3 <-  cut(ndns_1_11$Epcnt_4 ,
#                       breaks=c(0, 33, 45,63,80,100))
#ndns_1_11$UPF2 <- cut(ndns_1_11$Epcnt_4 ,
#                       breaks=c(0, 63,100))
ndns_1_11$zPF <-  ndns_1_11$NOVA_Epct_opti_4
tert_upf <- quantile(ndns_1_11$zPF ,probs = c(0,0.3,0.6,1), na.rm=TRUE)
                      
ndns_1_11$zPF3 <-  cut(ndns_1_11$NOVA_Epct_opti_4 ,
                       breaks=c(0, 33, 45,63,80,100))
ndns_1_11$zPF2 <- cut(ndns_1_11$NOVA_Epct_opti_4 ,
                       breaks=c(0, 63,100))
ndns_1_11$zPF4 <- cut(ndns_1_11$NOVA_Epct_opti_4 ,
                       breaks=tert_upf)

ndns_1_11$SYear <- as.numeric(ndns_1_11$SurveyYear)
ndns_1_11$ngor <- as.numeric(ndns_1_11$gor)

ndns_1_11$agegad3 <-  cut(ndns_1_11$age ,
                       breaks=c(0, 18,35,50,65,108))
ndns_1_11$hyp <- cut(ndns_1_11$omsysval ,
                       breaks=c(0,140,300))
ndns_1_11$BP <- ndns_1_11$omsysval 
ndns_1_11$Na <- ndns_1_11$Sodiummg
ndns_1_11$hiNa <- cut(ndns_1_11$Sodiummg , breaks = c(0,1500,3000,5000,6000,10000))
ndns_1_11$hiNa2 <- cut(ndns_1_11$Sodiummg , breaks = c(0,5000,10000))
ndns_1_11$BMI <- cut(ndns_1_11$bmival ,breaks = c(0,18,24,30,50))
```

```{r imd, echo=FALSE, message=FALSE, warning=FALSE}
#ndns_1_11$IMD <- na.omit(ndns_1_11[,c(,WIMD_2014_quintile,SIMD_2016_quintile,NIMD_2017_quintile)])
ndns_1_11 <- ndns_1_11[,IMD:=EIMD_2015_quintile]
ndns_1_11 <- ndns_1_11[is.na(IMD),IMD:=WIMD_2014_quintile]
ndns_1_11 <- ndns_1_11[is.na(IMD),IMD:=NIMD_2017_quintile]
ndns_1_11 <- ndns_1_11[is.na(IMD),IMD:=SIMD_2016_quintile] 
#ndns_1_11$IMD

```

```{r takinganymeds, echo=FALSE, message=FALSE, warning=FALSE,include=FALSE}


ndns_1_11$bpd <- if_else(ndns_1_11$diur2=="Taking drug", "Taking drug" ,
                  if_else(ndns_1_11$beta2=="Taking drug", "Taking drug",
                         if_else(ndns_1_11$obpdrug2=="Taking drug", "Taking drug", 
                                if_else(ndns_1_11$calciumb2=="Taking drug", "Taking drug", 
                                       if_else(ndns_1_11$calciumb2=="Taking drug", "Taking drug","not taking drug"), missing = "not taking drug"), missing = "not taking drug"), missing = "not taking drug"), missing = "not taking drug")
head(ndns_1_11$bpd)
```

```{r surveydesign,message=FALSE, warning=FALSE, include=FALSE}

ndns_1_11d <- svydesign(ids = ~area, weights = ~ wti_UKY1to11, strata=~astrata1, data = ndns_1_11, nest = TRUE)

#Exclusions
#medication affecting bp
ndns_1_11dd <- subset(ndns_1_11d, bpd == "not taking drug")
ndns_1_11f <- subset(ndns_1_11dd, Sex == "Female")
ndns_1_11na <- subset(ndns_1_11dd, Na >= 5000)
#ndns_1_11upf<- subset(ndns_1_11dd, UPF >= 63)
ndns_1_11zpf<- subset(ndns_1_11dd, NOVA_Epct_opti_4 >= 63)
ndns_1_11hyp <- subset(ndns_1_11dd, BP >= 140)
#ndns_1_11bp <- subset(ndns_1_11d, omsysval== is.na(NA))
#ndns_1_11bp

```

\newpage

# Results

## Participants

Considering participants who opted in and completed questionnaires, the whole NDNS population was 15,655.
The median age was 40.
Categorising age shows that 22% of the population was between 19 and 35.
There were 49% male participants.

After excluding those on medication, the population was 14217 participants.

This table [table 31](tab:table3) shows descriptive data.

Continuous variables are represented by the median and interquartile range in brackets.
Categorical variables give the number of participants and the percentage of the sample in brackets.

```{r table3, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE, paged.print=TRUE}
#| tab.id: "table3"
#| tbl-cap: "Characteristics of the sample population from National Dietary and Nutrition Study (2008-2019)"


keydatw1 <-tbl_svysummary(ndns_1_11d, include= c(Sex,Age, agegad3,   educfinh, IMD,region,SurveyYear ))
#keydatw1
keydatwdd <-tbl_svysummary(ndns_1_11dd, include= c(Sex,Age, agegad3, educfinh, IMD,region,SurveyYear ) )

keydatwzpf <-tbl_svysummary(ndns_1_11zpf, include= c(Sex,Age, agegad3,  educfinh, IMD,region,SurveyYear) )
keydatwna <-tbl_svysummary(ndns_1_11na, include= c(Sex,Age, agegad3,   educfinh, IMD,region,SurveyYear ) )
keydatwhyp <-tbl_svysummary(ndns_1_11hyp, include= c(Sex,Age, agegad3,    educfinh, IMD,region,SurveyYear) )

keydatwT <- tbl_merge(list(keydatw1,keydatwdd,keydatwzpf,keydatwna,keydatwhyp),tab_spanner = c("Whole Population", "Population not on BP medication","UPF >63%","Na >5000mg","hyp >140mmHg"))
table3 <- as_flex_table( keydatwT)
table3 <- set_caption(table3, "Characteristics of the Sample Population (National Dietary and Nutrition Study 2008-2019) ", 
  autonum = run_autonum(seq_id = "tab", bkm = "tab3"))
table3

#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/table1.png")
#save_as_image(width(autofit(keydatflx), width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/table1.png")
#save_as_docx(width(autofit(table3), width=5), path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/table1.docx")
#heatmap(keydataw1)
```

## Descriptive Data

The study population median age was 38.
The largest age group was 18-35.
49% of the participants were male.

The population exposed to UPF \>63% of their calories is made up of 4793 from a total of 14217 participants.
This compares with 9424 participants with lower exposure.
That is an exposure prevalence of 34%.

High UPF is more common in younger males than in the overall population or those not on medication.
37% are 0-16 years old.
There is a gradient in deprivation with more exposure in the most deprived group.
England's South East has 13% of those with high UPF intake.
With 11% in the North West.

The population exposed to Na \>5000mg has only 73 out of 14217 participants.
An exposure frequency of 0.5%.
61% are 18-35, and 80% male .
The least deprived makes up 5.2% of the participants compared with 39% of the most deprived.
The north has 46% of those with high Na intake, much the highest.

### Sensitivity to Survey Year

Internal consistency was examined by comparing background data across survey years.
This might also be called sensitivity analysis by survey year.
Wave 1 was a comparator for analysis of the other waves.

```{r tblVariablesbySurveyyear, echo=FALSE}
#| tab-id: "tbl-Key-Variables-by-Survey-year"
#| tbl-cap: "data quality continuous"

#ttestNa2 <- svyglm(Na~SurveyYear, ndns_1_11dd )
#ttestNa2t <- add_global_p(tbl_regression(ttestNa2))
#ttestEpcnt <- svyglm(NOVA_Epct_opti_4~SurveyYear, ndns_1_11dd )
#ttestEpcntt <- add_global_p(tbl_regression(ttestEpcnt))
#ttestOS <- svyglm(BP~SurveyYear, ndns_1_11dd)
#ttestOSt <- add_global_p(tbl_regression(ttestOS))
ttestage <- svyglm(Age ~ SurveyYear, ndns_1_11dd)
ttestages<-  add_global_p(tbl_regression(ttestage))
ttestbmi <- svyglm(bmival~SurveyYear, ndns_1_11dd)
ttestbmis <- add_global_p(tbl_regression(ttestbmi))
kvsearchyr <- as_flex_table(tbl_stack(list(ttestages,ttestbmis),group_header =  c("Age","BMI")))

kvsearchyr <- set_caption(kvsearchyr, "Sensitivity of Age and BMI by survey year from National Dietary and Nutrition Study (2008-2019) ", autonum = run_autonum(seq_id = "tab", bkm = "tab4"))
#kvsearchyr

#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/tbl-Variables-by-Survey-year.png")
#save_as_image(width(autofit(kvsearchyr), width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tbl-Variables-by-Survey-year.png")
#save_as_docx(width(autofit(kvsearchyr), width=5), path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tbl-Variables-by-Survey-year.docx")
```

The data had p.values \>0.05 for the controlled variables (age, sex, IMD) against annual wave.
UK region is part of the weighting, but this sample showed variation with p.value \<0.05.

[Table 4.2.3](tab:tbl-Categorical-variables-year) follows.

```{r tblCategoricalvariablesyear, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-Categorical-variables-year"
#| tbl-cap: "Key categorical variables by survey year from NDNS 2008-2018"

sexchi <- svychisq(~Sex +SurveyYear , ndns_1_11dd)
educhi <- svychisq(~educfinh+SurveyYear , ndns_1_11dd[,"SurveyYear" !=5])
imdchi <- svychisq(~IMD+SurveyYear , ndns_1_11dd)
vegchi <- svychisq(~vegetarn+ SurveyYear , ndns_1_11dd)
UPF3chi <- svychisq(~zPF3+ SurveyYear , ndns_1_11dd)
agegad3chi <- svychisq(~agegad3+ SurveyYear , ndns_1_11dd)
hypchi <- svychisq(~hyp+ SurveyYear , ndns_1_11dd)
BMIchi <- svychisq(~BMI+ SurveyYear, ndns_1_11dd)
gorchi <- svychisq(~region+ SurveyYear, ndns_1_11dd)
tablecat<- data.table("Variable"=c( "Sex", "IMD","Age","BMI","Region"),"p.value"= c(sexchi$p.value,imdchi$p.value,agegad3chi$p.value,BMIchi$p.value,gorchi$p.value ))
tablecat <- flextable(tablecat)
tablecat <- colformat_double(tablecat, digits = 2)

#tablecat <- footnote( tablecat, i = 1, j = 2,
#            value = as_paragraph(c("Chi Squared for categorical data")),
 #           ref_symbols = c("1"),
  #          part = "header")
#tablecat <- footnote( tablecat, i = 4, j = 1,
#            value = as_paragraph(c("year 5 data missing")),
#            ref_symbols = c("2"))
#tablecat <- footnote( tablecat, i = 6, j = 1,
 #           value = as_paragraph(c("year 11 data missing")),
 #           ref_symbols = c("3"))
tablecat <- valign(tablecat, valign = "bottom", part = "header")
tablecat <- autofit(tablecat)
tablecat <- set_caption(tablecat, "Sensitivity of categorical variables to survey year (NDNS 2008-2019) ", autonum = run_autonum(seq_id = "tab", bkm = "tab5"))
tablecat
#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/tablecat.png")
#save_as_image(width(autofit(tablecat), width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tablecat.png")
#save_as_docx(width(autofit(tablecat), width=5), path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tablecat.docx")
```

The exposure variables were compared across annual waves.
Most people have sodium exposure less than 3000mg.
In year one this was 81%, with only 0.9% exposed to more than 5000mg.
By year eleven 92% are reporting less than 3000mg, and 0.5% over 500mg.

UPF exposure was steady with 37%-40% of the participants exposed to 45%-63% throughout the survey.
Up to 38.2% of the participants were exposed to levels of more than 63%, the peak being in year 6.

```{r tblkeydata, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-keydata"
#| tbl-cap: "Exposure by survey year from National Dietary and Nutrition Study (2008-2019) "

keydatw <-tbl_svysummary(ndns_1_11dd,by =SurveyYear , include = c(hiNa,zPF3))
keydatw <- add_p(keydatw)
tblkeydata <-as_flex_table( keydatw)
tblkeydata <- set_caption(tblkeydata, "Exposure by survey year from National Dietary and Nutrition Study (2008-2019) ", autonum = run_autonum(seq_id = "tab", bkm = "tab6"))
#tblkeydata

#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/tblkeydata.png")
#save_as_image(width(tblkeydata, width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tblkeydatab.png")
#save_as_docx(width(tblkeydata, width=5), path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tblkeydata.docx")

```

Results were illustrated by plots against survey year, [figure 4](fig:fig-upf-and-survey-year) showed similarity between the waves for UPF intake and [figure 5](fig:fig-Na-and-survey-year) showed sodium exposure similarity between waves.

```{r figupfandsurveyyear, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-id: "fig-upf-and-survey-year"
#| fig-cap: "Energy from UPF% in each annual cohorts NDNS (2008-2019)"


svyboxplot(zPF~SurveyYear,ndns_1_11dd, na.rm = TRUE,main = "UPF% by survey year", ylab = "UPF%" ,xlab ="Survey Year")
```

```{r figNaandsurveyyear, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-id: "fig-Na-and-survey-year"
#| fig-cap: "Na in mg in each annual cohort NDNS(2008-2019)"
svyboxplot(Na~SurveyYear,ndns_1_11dd,main = "Sodium Intake in mg by survey year", ylab = "Sodium intake in mg" ,xlab ="Survey Year")

```

## Outcome variable

```{r figBPandsurveyyear, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-id: "fig-BP-and-survey-year"
#| fig-cap: "Plot of the BP in mmHg by year from NDNS (2008-2018)"

svyplot(BP~SurveyYear,ndns_1_11dd, na.rm = TRUE,main = "BP in mmHg by survey year", ylab = "BP in mmHg" ,xlab ="Survey Year")
```

[figure 6](fig:fig-BP-and-survey-year) shows that the mean BP is consistent across the waves.

The population with BP \> 140 mmHg is 876 participants.
This gives a prevalence of 6%.Men are once again overrepresented.
These participants are older than the population median, 44% of these participants are over 65.
This group are statistically significantly different from the populations with high Na, or high UPF.
There is a reverse gradient with IMD in this population.
The most deprived are least represented in this population.
The largest proportion are in the least deprived category.
The north is second highest in raised BP after the south east.
The BP was highest in year one with 125 mmHg, and the lowest 120 mmHg in year 6.
BP rose through life to a mean of 134 mm Hg in the over 65 age category.

```{r tblbpdata, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-bpdata"
#| tbl-cap: "BP by survey year from National Dietary and Nutrition Study (2008-2019) "

tblbpdata <-tbl_svysummary(ndns_1_11dd,by =SurveyYear ,include = c(hyp))
tblbpdata <- add_p(tblbpdata)
tblbpdata <-as_flex_table( tblbpdata)
tblbpdata <- set_caption(tblbpdata, "BP by survey year from National Dietary and Nutrition Study (2008-2019) ", autonum = run_autonum(seq_id = "tab", bkm = "tab7"))
#tblbpdata

#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/tblkeydata.png")
#save_as_image(width(tblkeydata, width=5), "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tblkeydatab.png")
#save_as_docx(width(tblkeydata, width=5), path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/tblkeydata.docx")

```

## Main Results

The main results are the correlation between the exposure and the outcome variables.
Univariable regression demonstrates the interaction with other explanatory variables.
Then multivariable regression is used.
Mathematical models containing explanatory variables are constructed and compared using 'goodness of fit' statistics.

In the simplest model, by calculation using a Chi squared 2\*2 table, the odds ratio for hypertension in participants exposed to UPF \>63% is 0.5.
The odds ratio for hypertension in participants exposed to Na \>5000mg is 1.45.
These results take no account of weighting, or confounding.

```{r chisq, echo=FALSE, message=FALSE, warning=FALSE}
hyp1chi <- svychisq(~hyp+ hiNa2 , ndns_1_11dd, na.rm=TRUE)
#hyp1chi$p.value
#round(ftable(hyp1chi$observed),2)


hyp2chi <- svychisq(~hyp+ zPF2, ndns_1_11dd, na.rm=TRUE)
#hyp2chi$p.value
#round(ftable(hyp2chi$observed),2)
```

```{r tblunivariableregressionsb, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-univariable-regressionsb"
#| tbl-cap: "regressions in pairs between key variables full population showing direction and strength of relationships NDNS data 2008-2019"
#univariable regression models
#UPF Energy % only
lm1E <- svyglm(BP ~ zPF, ndns_1_11dd)
lm1Et <- tbl_regression(lm1E)
#univariable regression models
#sodium only
lm1na <- svyglm(BP ~ Na, ndns_1_11dd)
lm1nat <- tbl_regression(lm1na)

lmEna <- svyglm(zPF ~ Na, ndns_1_11dd)
lmEnat <- tbl_regression(lmEna)

lmA <- svyglm(BP ~ Age, ndns_1_11dd)
lmAt <- tbl_regression(lmA)

lmEA <- svyglm(zPF ~ Age, ndns_1_11dd)
lmEAt <- tbl_regression(lmEA)

lmNaA <- svyglm(Na ~ Age, ndns_1_11dd)
lmNaAt <- tbl_regression(lmNaA)

#row1<- tbl_merge(list(lm1nat,lm1Et,lmEnat),tab_spanner = c("BP","UPF","Na"))
#row2 <- tbl_merge(list(lmAt,lmEAt,lmNaAt),tab_spanner = c("BP","UPF","Na"))
#as_flex_table(tbl_stack(list(lm1nat,lmEnat,lm1Et,lmEAt,lmAt,lmNaAt),group_header = c("BP/Na","UPF/Na","UPF/bp","UPF/Age","Age/BP","Age/Na")))
```

Univariable regression is adjusted for weighted survey samples.
Identified important relationships within the data.
Confounding plays a part in these results.

The result for Sodium against UPF shows that there is no linear relationship between Sodium and UPF, in this table [Table 4.4.1](tab:tbl-univariable-regressions).
UPF compared to Na also shows a zero beta value indicating no linear relationship.

UPF does show a negative relationship with BP, which is statistically significant, beta = -0.19 (CI -0.22,-0.15).

There is also a negative relationship for UPF with Age, again statistically significant beta = -0.25 (CI -0.26,-0.23).

Age has a relationship with BP with a statistically significant positive gradient beta= 0.43 (CI 0.41,0.45).

There is also a positive relationship with Na, which is also statistically significant 1.5 (CI 0.77,2.3).

```{r tblunivariableregressions, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-more-univariable-regressions"
#| tbl-cap: "Univariable Regression NDNS data 2008-2019"
#univariable regression models
#UPF Energy % only
lm1bmi <- svyglm(BP ~ bmival, ndns_1_11dd)
lm1bmit <- tbl_regression(lm1bmi)
#univariable regression models
#sodium only
lm1Ag1 <- svyglm(BP ~ agegad3, ndns_1_11dd)
lm1Ag1t <- tbl_regression(lm1Ag1)

lmEbmi <- svyglm(zPF ~ bmival, ndns_1_11dd)
lmEbmit <- tbl_regression(lmEbmi)

#lmed <- svyglm(BP ~ educfinh, ndns_1_11dd)
#lmedt <- tbl_regression(lmed)

lmEs <- svyglm(zPF ~ agegad3, ndns_1_11dd)
lmEst <- tbl_regression(lmEs)

#lmEed <- svyglm(zPF ~ educfinh, ndns_1_11dd)
#lmEedt <- tbl_regression(lmEed)

lmNabm <- svyglm(Na ~ bmival, ndns_1_11dd)
lmNabmt <- tbl_regression(lmNabm)

#lmNaed <- svyglm(Na ~ educfinh, ndns_1_11dd)
#lmNaedt <- tbl_regression(lmNaed)

lmNaAg <- svyglm(Na ~ agegad3, ndns_1_11dd)
lmNaAgt <- tbl_regression(lmNaAg)

lmNaI <- svyglm(Na ~ IMD, ndns_1_11dd)
lmNaIt <- tbl_regression(lmNaI)

lmzPFI <- svyglm(zPF ~ IMD, ndns_1_11dd)
lmzPFIt <- tbl_regression(lmzPFI)

lmBPI <- svyglm(BP ~ IMD, ndns_1_11dd)
lmBPIt <- tbl_regression(lmBPI)
#row1<- tbl_merge(list(lm1nat,lm1Et,lmEnat),tab_spanner = c("BP","UPF","Na"))
#row2 <- tbl_merge(list(lmAt,lmEAt,lmNaAt),tab_spanner = c("BP","UPF","Na"))
univariablesUPF <-tbl_stack(list(lmEnat,lmEAt,lmEst,lmEbmit,lmzPFIt) )
univariablesNa <- tbl_stack(list(lmNaAt,lmNaAgt,lmNabmt, lmNaIt))
univariablesBP <- tbl_stack(list(lm1Et,lm1nat,lmAt,lm1Ag1t,lm1bmit, lmBPIt))

univariables2 <- tbl_merge(list(univariablesBP,univariablesUPF,univariablesNa),tab_spanner = c("**BP**","**UPF**", "**Sodium**"))
univariables2 <- as_flex_table(univariables2)
#univariables <- as_flex_table(tbl_stack(list(lm1nat,lmEnat,lm1Et,lmEAt,lmAt,lmNaAt,lm1bmit,lm1Ag1t,lmedt,lmEbmit,lmEst,lmEedt,lmNabmt,lmNaAgt,lmNaedt),group_header = c("BP/Na","UPF/Na","UPF/bp","UPF/Age","Age/BP","Age/Na","BP/bmi","BP/Agg1","BP/ed","UPF/bmi","UPF/age","UPF/ed","Na/bmi","Na/Agg","Na/ed")))
univariables2<- set_caption(univariables2, "Univariable Regression (NDNS data 2008-2019)", autonum = run_autonum(seq_id = "tab", bkm = "tab8"))
univariables2
#png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/univariables.png")
#save_as_image(univariables, width=5, "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/univariables.png")
#save_as_docx(univariables, width=5, path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/univariables.docx")
```

Multivariable regression models were constructed.
They are regressed against hypertension in patients who are not on BP reducing medication.

The model, "Sodium Only", includes sodium as the exposure variable.
The odds ratio for the group taking between 5000mg and 6000mg per day is statistically significantly different from those taking less than 3000mg per day.
There is an odds ratio of 5.20 (CI 1.39,19.5) for this group.

"UPF only" shows a significant difference in odds ratio 0.60(CI 0.36,0.99) for the group 63-80%.

The last model, "Sodium and UPF", shows that when combined the effect remains.
The odds ratio for 5000-6000mg of Na remains statistically significant 5.57(1.47,21.2).
The odds ratio for UPF also remains 0.57(0.34,0.94).
These are both changed from the separate models.
The Akaike Information Criterion (AIC), a measure of 'goodness of fit', is lower for this combined model, 3590.81, indicating it is a better fit for the data also.

[Table 4.5.1](tab:tbl-multivariable-outputs-bp) follows below.

```{r bpmultivariablemodels, echo=FALSE, warning=FALSE}

lm1AS <- svyglm(hyp ~  Sex +BMI +agegad3+educfinh + IMD +SurveyYear,family = binomial(link = "logit"), ndns_1_11dd , na.action = na.exclude)

theme_gtsummary_mean_sd()

#sodium only
lmna <- svyglm( hyp ~ hiNa  +Sex +agegad3 + BMI+educfinh+  IMD+SurveyYear ,family = binomial(link = "logit"), ndns_1_11dd,na.action = na.exclude)

lmnaE <- svyglm(hyp ~ hiNa +Sex +agegad3+ zPF3 + BMI +educfinh+ IMD +SurveyYear,family = binomial(link = "logit"), ndns_1_11dd,na.action = na.exclude)

lmME <- svyglm( hyp~ zPF3 +agegad3+Sex + BMI +educfinh + IMD +SurveyYear,family = binomial(link = "logit"), ndns_1_11dd,na.action =na.exclude)

#lmME2 <- svy_vglm(UPF3 ~ hyp+hiNa +agegad3+Sex + bmival +educfinh + EIMD_2010_quintile +bpd,family = multinomial(refLevel=1),design= ndns_1_11d,na.action =na.exclude)
#tbl_regression(lmME2)

```

```{r tblmultivariableoutputsbp, echo=FALSE, message=FALSE, warning=FALSE}
#| tab-id: "tbl-multivariable-outputs-bp"
#| tbl-cap: "Table of multivariable regression against BP to identify the effects relating to Na and UPF NDNS data 2008-2019"
#lm1ASt <-tbl_regression(lm1AS,include=c(hiNa,NOVA_Epct_opti_4) ,exponentiate = TRUE)
lmnat <- tbl_regression(lmna,include=hiNa,exponentiate = TRUE, label = list( hiNa ~ "Sodium Intake mg"))
#lmnat$model_obj$R
#lmnat <- add_glance_table(lmnat,include = c( AIC))
lmnaEt <- tbl_regression(lmnaE,include=c(hiNa,zPF3),exponentiate = TRUE, label = list(zPF3 ~  "Ultraprocessed Food %", hiNa ~ "Sodium Intake mg"))
#lmnaEt <- add_glance_table(lmnaEt,include = c( AIC))
lmMEt <- tbl_regression(lmME,include=zPF3,exponentiate = TRUE, label = list(zPF3 ~ "Ultraprocessed Food %"))
#lmMEt <- add_glance_table(lmMEt,include = c( AIC))

multivariables <- tbl_merge(list(lmnat,lmMEt,lmnaEt), tab_spanner = c("Na only","UPF only","Na and UPF"))
#multivariables <- tbl_stack(multivariables)    
multivariables <- as_flex_table( multivariables )
multivariables<- set_caption(multivariables , caption = "Table of multivariable regression against BP to identify the effects relating to Na and UPF NDNS data 2008-2019", autonum = run_autonum(seq_id = "tab", bkm = "tab9"))

 #multivariables <- tbl_stack(multivariables,AICt)
multivariables <- add_footer_row(multivariables, values = c("AIC",round(lmna$aic,2),round(lmME$aic,2),round(lmnaE$aic,2)), colwidths = c(1,3,3,3))
multivariables <- add_footer_lines(multivariables, c("All models include additional variables Sex,Age, BMI, Education, IMD and Survey Year"))

# AICt <- flextable(data.table("AIC" =c( lmna$aic,lmME$aic,lmnaE$aic)))                                  
                                   
multivariables
# png("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/multivariables.png")
#save_as_image(multivariables, width=5, "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/multivariablesb.png")
#save_as_docx(multivariables, width=5, path = "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/multivariables.docx")
```

### Relative Effect SIze calculation

Using the AIC statistic assessing goodness of fit for each model gives another way of understanding the comparative effects between variables.
The lowest scored model is the optimal model.
The 'best' of these models is that with only sodium included "Na only".
The UPF models both being further away from the lowest value.

Of the difference between the lowest scoring model and the highest 70/30 is due to the sodium.
the additional effect of UPF is in proportion to this.

\newpage

# Discussion

## Key Results

This analysis shows a statistically significant correlation between high Na intake and hypertension .
There was an increased odds ratio of hypertension with higher sodium intake (OR=5.57(1.47,21.2)).
Between high UPF and hypertension there was a lower odds ratio of hypertension with high upf intake (OR=0.57(0.34,0.94)).
This is present in a combined multivariable logistic regression model.
This technique helps control explanatory variables.

The effect is also present in the multivariable logistic regression model of each individual variable, for sodium this is OR=5.20(CI 1.39,19.5), and for UPF OR=0.60(CI 0.36,0.99).

Sex and age contribute the largest effects.

In univariable analysis with no controlling for confounding or explanatory variables, there was no correlation between UPF intake and sodium intake (beta=0).
There was a strong correlation between age and BP (beta =0.43 (CI 0.41,0.45)), as well as age and Sodium intake.
There was a strong age gradient of UPF intake (beta= -0.25 (CI -0.26,-0.23)).

The descriptive analysis does show the size of different groups.
It also shows how they differ from the overall population.
The size of some of these subgroups is a reason for caution over the overall results.

## Limitations

This is a cross-sectional study, and so has the limitations of this design.
In particular, exposure and outcome are measured at the same time.
Causal relationships cannot be understood from this data.

Time and age have particular effects in cross sectional studies.
Prospective and longitudinal studies, such as cohort or RCT studies have less of this affect.
This sometimes results in inverting relationships between variables.

This study was organised by government departments connected with food and farming alongside the Department of Health.
The sample was designed to monitor relevant outputs.
Funding and commissioning processes affect design structure and might also affect participant engagement and expectation.
The study was not powered sufficiently for in-depth subgroup analysis.
It was powered for monitoring of food intake across the UK population.

### Bias

Selection bias was approached by using random selection of participants using a carefully constructed stratification model.
Addresses were selected by postal units to ensure geographic spread of participants.
This ensured that whilst random the sample remained representative.

Uptake and Drop out bias was approached by ensuring that sample sizing included scope for this to enable comparable sample sizes across annual waves.

Social desirability bias acknowledges that participants remember and record intake framed by their beliefs about the needs of the study, and their beliefs about what is perceived as being healthy.
To examine this, in the first wave, Lennox et al @lennox conducted a double labelled water study.
They compared reported energy intake with measured values.
They showed some significant differences between measured energy intake, and reported energy intake with differences between different age groups.

At the analysis stage weighting was used to standardise the sample for several variables.
Those selected were Age, Sex, region and IMD.
Weights are available for different levels of analysis as participants who did not complete the initial interview were not selected for subsequent blood analysis.

Sensitivity analysis looked to see how the variables changed over the survey years after weighting.
This is intended to assess changes in sampling over the course of the study.

The result depends on participants recording foods in the same way as time goes on.
Exposure of the whole population to a stimulus to change their diet or the recording of their diet may result in systematic changes in results.

In years 9-11 a slight difference for identifying foods for analysis as researchers have started to become aware of the need to understand 'processing'.
This may account for the apparent lower exposure in the last three years.

These changes would affect the outcome variable less.
However BP measurement technology has changed over ten years.
BP machines derive their results from the changes in pressure detected in the arm of the participant, the algorithm used by the sphygmomanometer may have changed.

Weighting maintains age, sex, IMD, and government region across the waves.
BMI is no different, and educational attainment is also unaffected.

The populations do change over time as some of the added variables do show statistical significant changes.

In populations with exclusions the careful sample selection and weighting are overcome by the biasing effect of different selections.
Selecting for BP, UPF or sodium changes the cohort sex balance, age range and IMD pattern making these samples less representative.

## Interpretation

Participants with high Na \>5000mg are more likely to have hypertension.
He @hefjm , Graudal @graudal2014 , Strazzullo @strazzullo2009 , inform us that high Na intake is contributing to high BP.
However, causation cannot be identified in this study.
This seems to be the larger effect with the change in AIC being greater than the change for UPF.

Low UPF intake is correlated with high BP.
Only correlation is certain, and reverse correlation may be an issue.
Wang @wangAssociationUltraprocessedFood2022 , Barbosa @barbosa2022 , and Mambrini @mambrini2023 all find a positive correlation using longitudinal studies such as those of Scaranni @ScaranniPatriciadeOliveiradaSilva2021Ufci and Mendonça @mendoncaUltraProcessedFoodConsumption2017 .
The descriptive data shows that those with \>63% UPF are a peculiar subgroup.
They are different in age from those with high sodium or high BP.

### High UPF intake

Mertens @mertens2022 identifies that the UK has one of the highest % intake of UPF in Europe.
The USA, Canada and Australia have similarly high levels.
Other European countries are still fighting to retain a different food culture, Touvier et al with Nutrinet @srour2019 @touvier2020 .

Colombet @ColombetZoé2021Dass highlights countries in the rest of the world at differing levels of 'nutrition transition'.
This might be influenced by the degree of 'westernisation'/'internationalization'/ or 'capitalist colonialisation' into local culture, as well as by more general socio-economic factors @colombet2022 .

### UPF and sodium

One of the odd findings is that there is no relationship between % UPF and sodium intake.
Webster et al @websterSystematicSurveySodium2010 in Australia, and Ni Murchu @nimhurchu2011 looking at 44,000 foods in the UK, show that UPF are high in salt.
If UPF is 'high in salt' then high UPF should be correlated with sodium and BP.
This contradictory finding that there is no correlation suggests that UPF varies in quality.

The highest salt intakes are amongst the group with the highest UPF intakes, but there may be a qualitative difference in the intake.
Rauber et al @rauber2019 and @rauber2020 using this dataset identified high free sugar intake leading to obesity and type 2 diabetes, from drinking large quantities of artificially sweetened beverages.
As Hall @hall2019 demonstrated in a randomised controlled trial, UPF intake is affecting the choice to eat.
Therefore, it is not as simple as high UPF leads to high sodium leads to hypertension.
Hence, simple reformulation, to reduce salt intake, will not be effective.

The effect of sodium is augmented by that of UPF.
The two together is greater than either individually

### A synergistic effect

The regression model including both exposure variables had a better fit of data than either individually.
This finding suggests that the effect of each variable is not the whole effect, but that there is a synergistic effect.
This would fit with the idea that Monteiro's concept of UPF @monteiro2010 is wider than being simply a nutritional effect.

This effect might be due to broader biologic effects, structural, or energy density as Rauber explains @rauber2020 .
Alternatively, it could relate to the non-biologic, the wider economic and cultural aspects of UPF.

Bourdieu @bourdieu2002 studied how food and culture are intertwined.
Humphries' @humphries2016 editorial and Martinez's @martínez2021 deconstruction update these wider aspects with international and epistemological distinctions.
These explanatory papers help to develop our understanding and provide a framework for further study.

### Age

Age is a particular feature in this outcome.
BP is very strongly affected by age.
In cross sectional studies Age has several dimensions.
Age identifies cohorts of people with particular experiences, it identifies duration of experience, it represents physiologically different states, and it also identifies access to resources financial, material, and experiential.
Prospective, longitudinal studies, and case matching can help reduce some of the effects.
However longitudinal studies have reported similarly equivocal results, identifying a potentially more complex interaction.

### Ideas for further research

Further research based on the findings is needed for confirmation by further review and analysis.
The finding of a synergistic effect, and to understand the types of UPF ingested in this population would be priorities.

This further study might also include attempts to better map Martinez's putative epistemological framework.
This could be by looking quantitatively for more specific economic and cultural markers in the data.
Qualitative approaches might also be engaged, perhaps in a mixed methods approach to identify the importance of the findings to real populations in a local context.

#### Public Health Policy ideas

This study aims to inform local policy to reduce BP and so Non-Communicable Disease.
If UPF and sodium intake increase the risk of hypertension then policy to reduce exposure might deliver change at a population level.
This study supports the case for a place based food strategy approach @liverpoo2018 .

Policy is an 'upstream' approach.
Katikreddi @katikireddi2021 recently showed how Diderichson's model @diderichsen2019 for policy intervention can be used to understand exposure indirectly or directly when there are health inequalities.

Reformulation is a policy suggestion, where UPF is further processed to remove the salt.
By demonstrating that UPF is not just high in salt this study supports the argument of avoiding further formulation as a policy for reducing BP.
Pearson-Stuttard et al @pearson-stuttard2018 model the effects of salt reformulation, and identify economic and social benefits.
They make no mention of negative effects of UPF that might not be due to salt, which might reduce the overall benefit of their model.

Dietary approaches to improving public health are able to deliver proportionate and universal interventions to populations @marmot2010 to reduce the incidence of NCD.
When delivered up stream at the policy level they are effective and efficient and minimise cost.
These approaches offer significant benefits over actions targeted at individuals.

Dietary and 'awareness' approaches can be used by individuals.
These approaches risk the development of a culture of blame of individuals and of sub-groups in society.
The commercial and social determinants of health @diderichsen2001 @whitehead1988 play out a significant role in research, and delivery of public health improvements around food.

Dimbleby @dimbleby proposes a much broader strategic approach to the whole economy of food.
Tulleken @tulleken2023 identifies that we need greater care in licencing ingredients, and helps health promotion.
People who know that UPF is bad, are more likely to accept policy limiting availability.

## Generalisability

This study used national data.
This was stratified across the four home nations.
It was stratified for IMD, and sex and to cover adults and children.
The study can therefore be generalised to the UK population.
The results are comparable with those in Korea, Brazil and USA.
These include countries with lower UPF intake, but also similar levels.

Caution is in order when extrapolating down to local areas from national data.
Age and sex standardisation might make the datasets more similar.
IMD differences are harder to control.

\newpage

# Conclusion

In summary,

In this representative sample of adults and children across the four home nations of the UK between 2008 and 2019, exposure to higher sodium dietary intake, and lower UPF dietary intake, increased the odds of having a mean systolic blood pressure of over 140mmHg.

Combining the two exposures had a larger than expected effect on the 'goodness of fit' of the model.
This suggests a broader effect than nutrition alone, possibly measuring some of the economic and cultural aspects of UPF.

UPF intake in this study was not correlated with high sodium intake, suggesting caution around models proposing reformulation as an effective approach to reducing BP.

### Further Recommendations

Further quantitative and qualitative research will be needed to understand this result.

A strategic approach to food policy might be needed, independent of reformulation.

\newpage

# Bibliography

::: {#refs} :::

# Appendix {.unnumbered}

# Appendix 1 Approved Proposal

The approved proposal

# Appendix 2 Ethics Certificate

The ethics cert.

# Appendix 3 Software used

The software used

CRAN

GT Summary @gtsummary

```{r packagecite, echo=FALSE, message=FALSE, warning=FALSE, warnings=FALSE}
#cite_packages(output="citekeys", out.dir = getwd() ,out.format= "Rmd")

```
