---
title: "bp upf and na"
output:
  word_document: 
    toc: yes
    fig_caption: yes
    keep_md: yes
  html_document:
    toc: yes
    fig_caption: yes
    number_sections: yes
    keep_md: yes
  pdf_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: sentence
---


 
# BP and UPF and Na in NDNS Dissertation
# Method
## Introduction
This section will take the research question and explain how the data will be used to answer the question.

There will be a description of the study and how the data collection happened.
This will be followed by a consideration of the process governing ethics in this research.

The data will be loaded and the relevant variables identified and extracted. 
Some data may need to be recalculated or to be processed to make a more useable form.
The population will be reviewed and any groups which might influence the results removed.
Once the data set has been prepared, analysis begins with description of the data.
The second analysis section involves using linear regression to identify if there is a correlation
between the BP and each of the key variables. 
Multivariable regression models are then generated.
These models are finally examined to identify the relative importance of the different variables in 
developing an optimal model. 


## Research Question

What proportion of the association between blood pressure (SBP/DBP) and UPF intake can be explained by the changes in salt intake in England between 2008-2012 and 2015-2019? 

The question can be split into parts, 
What was intake of salt between 2008 and 2019?
What was intake of UPF between 2008 and 2019?
What was BP between 2008 and 2019?
Did each of these change over that time and how?
Did the changes in any one affect any other?
What are the sizes of the changes?

All of these questions look for numbers as answers. 

Answering the question starts with counting.

## National Dietary and Nutritional Survey

This survey is a collaboration between government departments responsible for health and for food production.
They have engaged academic partners to deliver reports on diet and nutrition across the United Kingdom.
The study is designed to be representative across the whole area.

### Study design

This is a rolling cohort study which each year selects a new cohort of participants. 
The sample is approximately 1000 per year with 50% adults. 
The design has a random selection across postal units (psu). 

Having taken up the study participants complete a 4 day food diary, and have an interview with a nurse which includes taking several measurements.

### NDNS Dataset

The data from the NDNS study contains items about each individual,and their household.
It contains a table with each item of food as recorded in their diary.
There is a table with the overall intake of each of a large range of nutrients for the whole period.
This is calculated from the diary using nutritional tables which are published as part of the dataset.
The dataset is available via the UK national Data service for research purposes.


NDNS began before Monteiro's processing based classification was developed. 
There is no record of Nova food type in NDNS. This has bee calculated from the food descriptions. 
I have used a table from Rauber et al., but also one from Colombet (personal communication)

### University Research Governance and Ethical Review

The research has beencarried out under the Universtiy governance. A proposal was discussed and agreed within the department. 
The need for ethical review was considered using the university research tool. 
The fact that the data is anonymised and there was no contact with participants means that there is minimal risk of hram to research participants.

Other ethical issues include data custodianship ensurign that the the rights of the owners of the data and of the participants are still considered as part of the process of analysis and dissemination of the research.

Issues around the power structures which lead to privelage one research project or proposal over another are considered more in the positionality section.

## Data Processing

The storage of the data is in keeping with the research governance agreements of the University and the Data set owners.
The data is read from its files using 'r-studio' with the processing being carried out using packages available from CRAN.
I have used files which had been amalgamated into four batches. These are 2008-2012, 2013-2014, 2015-2016, 2017-2019.

Once the data labels are made consistent across the batches, weighting recalculation is done.
This generates values which account for differences in population balance across the annual cohorts. 
These result from differences in compliance and uptake within and across the years.

The years are amalgamated and the nature of the variables is specified.

The food diaries need processing to identifiy the UPF intake.
Each persons food diary entries are assessed against the Nova food classification from Rauber. 
Then the weight and energy content of the days food is calculated by Nova group. 
This is added to the intake for the other 3 days and the total intake by nova group established.

The percentage of the total intake by weight and by energy is then calculated for each of the 4 nova categories.
Nova group 4 or UPF intake is used for the study.



## Method for Data Analysis

### Descriptive data analysis


The data is summarised, with Mean median, and range for the key continuous variables.
The key variables are omsysval, UPF intake and sodiummg.
These variables are the ones which most relate to the research question.
First for years 1-4 then for 9-11.

There are a number of related variables in the dataset.
These were chosen for reliability and practicality.
These variables are ones which can also influence BP.
They include Age, Sex, BMI, height and weight.
The population for years 1-4 are compared with those for years 9-11.

The omsysval is a validated measurement with significant quality assessment within the dataset.
Raw systolic values are present in the dataset but are made up of data with issues around quality.
In particular the systolic values are assessed for the effects of exercise, temperature and ill health.
The variable omsysval is a quality assured mean value which is reliable across the dataset.

The sodium value is one calculated from intake based on food diaries and standard food nutrient values.
This only reflects standard foods and is the result of assumptions about the content being consistent.
Serum sodium values are available for the early dataset, but not the later one.
There are also values for 24 urinary sodium which is probably a better indicator of dietary sodium for parts of the dataset, but again these are not found in both time periods.
The Data is displayed in table x with minimum, mean, median, and maximum values. 
The exposure variables are sodium intake (Sodiummg), and ultra processed food intake (UPF_4).
The outcome variable are the mean systolic blood pressure (omsysval). 


Key additional variables are considered in the later section looking at how they change across the populations.
Age, sex, and bmi are important contributors. Education, NSSEC and IMD are also looked at.

### Exclusions 

The relationship between salt and systolic blood pressure may be different in individuals with pathologically high BP. 
Those taking BP controlling medications may have a different relationship to sodium and UPF.
These patients were excluded from the main analysis.
Analysis was done with them included and this produced results in line with those presented, but of greater magnitude.
This additional analysis is not presented here.

### Comparative Analysis

The second phase of analysis uses standard techniques to compare the means of the populations. 
A student's t test is used as the populations are normally distributed continuous variables. 
The t tests compare the sodium, UPF, and systolic BP in the cohort from 2008-2012 with that from 2017-2019.
This will show how the intake has changed, and the same for the outcome.
These are not the same participants so matched analysis, or time series analysis is not directly applicable.

Plots will be given to show the values in each of the available eleven cohorts.

Other variables in the data are compared across to assess how the data changes. 
t tests are again used for continuous data, and chi squared tests are used for categorical data.

### Linear Regression

Analysis of the correlation between BP and sodium intake, and then UPF intake is done using linear regression. 
This will give an indicator of the direction, and strength of any relationship between the variables.
There is also anova analysis to understand the statistical significance of these results.

### Multiple Regression

Multivariable regression models are then developed to understand the interactions between variables and to develop a mathematical model. 
The optimal model is one which best explains the pattern of data, but which also makes practical sense for the wider understanding of relationships.
Assessment techniques try to understand the importance of including particular variables, and the form in which they are best included. 

### AIC and sensitivty Anaylsis

This section compares models side by side using assessment techniques to identify the best way of describing the data. 
The 'best' in part is determined by the whether a mode is needed to predict more data, or just to understand the data available.
Here it is about how best to describe the relationship between Na, UPF, and BP.

##  Method Conclusion

This section has highlighted how the material for the study is brought together and how the governance and ethics fit with the data collection, processing and analysis to help us to derive the results which will be presented in the next section.




















```{r include=FALSE}
library(haven)
library(data.table)
library(readxl)
library(sensemakr)
library(labelled)
library(survey)
library(srvyr)
library(gtsummary)
```

```{r read data, message=FALSE, warning=FALSE, include=FALSE}

sav4 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_personleveldietarydata_uk_v2.sav"

sav4b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_indiv_uk.sav"

sav6 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_personleveldietarydata_v2.sav"

sav6b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_indiv.sav"

sav8 <-"/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_personleveldietarydata.sav"

sav8b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_indiv.sav"



sav11 <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_personleveldietarydata_uk_20210831.sav"

sav11b <- "/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/upfnabp/data/UKDA-6533-spss/spss/spss25/ndns_rp_yr9-11a_indiv_20211020.sav"

sav4r <- as.data.table(read_sav(sav4))
sav4br <-as.data.table(read_sav(sav4b))

sav6r <- as.data.table(read_sav(sav6))
sav6br <-as.data.table(read_sav(sav6b))

sav8r <- as.data.table(read_sav(sav8))
sav8br <-as.data.table(read_sav(sav8b))


sav11r <- as.data.table(read_sav(sav11))
sav11br <- as.data.table(read_sav(sav11b))
sav11br$area <- sav11br$Area
```

```{r std names4, message=FALSE, warning=FALSE, include=FALSE}
#standardise the variable names for grp 1-4 
sav4br$bpmedc2 <- sav4br$bpmedc
sav4br$bpmedd2 <- sav4br$bpmedd
sav4br$diur2 <- sav4br$diur
sav4br$beta2 <- sav4br$beta
sav4br$aceinh2 <- sav4br$aceinh
sav4br$calciumb2 <- sav4br$calciumb
sav4br$obpdrug2 <- sav4br$obpdrug

sav4br$ethgrp5 <- sav4br$ethgr5
sav4br$ethgrp2 <- sav4br$ethgr2

sav4br$hyper140_2 <- sav4br$hyper140
sav4br$hibp140_2 <- sav4br$hibp140
sav4br$hyper1_2 <- sav4br$hyper1
sav4br$highbp1_2 <- sav4br$highbp1

sav4br$educfinh <- sav4br$EducFin
sav6r$SurveyYear <- sav6r$Surveyyear
```

```{r stdnames11, message=FALSE, warning=FALSE, include=FALSE}
sav11br$htval <- sav11br$htval2
sav11br$wtval <- sav11br$wtval2
sav11br$bmival <- sav11br$bmival2

sav11br$gor <- sav11br$GOR
sav11br$age <- sav11br$AgeR
sav11r$Age <- sav11r$AgeR
```

```{r novafile, message=FALSE, warning=FALSE, include=FALSE }
NOVA_ForDavid <- as.data.table(read_excel("data/NOVA_ForDavid.xlsx"))

NOVA_ForDavid[, Sex := factor(Sex, levels = 1:2)]
NOVA_ForDavid[, Country := factor(Country)]
NOVA_ForDavid[, SurveyYear := factor(SurveyYear)]
NOVA_ForDavid[, Age := as.numeric(Age)]
```

```{r weight, echo=FALSE}
#Doing the weighting correction, as explain in the PDF "6533_ndns_yrs9-11_weights_guide"

n1 = sum(sav4br$wti_UKY1234)  #Should be 6828

n2 = sum(sav6br$wti_Y56)         #Should be 2546

n3 = sum(sav8br$wti_Y78)         #Should be 2723

n4 = sum(sav11br$wti_Y911)       #Should be 3558

N <- sum(n1, n2, n3, n4) #Should be 15655
```



```{r message=FALSE, warning=FALSE, include=FALSE}
#now the proportion weight for each is calculated
sav4br$wti_UKY1to11 <- ((sav4br$wti_UKY1234/n1) * N)  * (4/11)

sav6br$wti_UKY1to11 <- ((sav6br$wti_Y56/n2) * N) * (2/11)

sav8br$wti_UKY1to11 <- ((sav8br$wti_Y78/n3) * N) * (2/11)

sav11br$wti_UKY1to11 <- ((sav11br$wti_Y911/n4) * N) * (3/11)

```






```{r allsetup variables, message=FALSE, warning=FALSE, include=FALSE}
 #measured and recorded data yr1-4a_indiv_uk 
#subsets of the table to identify grouped information
bpset <- c("seriali","omsysval","omdiaval")
ethnset <- c("ethgrp5","ethgrp2")
saltset <- c("SaltChk","SalHowC","SltSHow" ,"vegetarn")
medsset <- c( "bpmedc2","bpmedd2","diur2","beta2","aceinh2", "calciumb2","obpdrug2", "PregNowB")
hypset <- c("hyper140_2", "hibp140_2", "hyper1_2", "highbp1_2")
incset <-c("nssec8","gor","region", "EIMD_2007_quintile","EIMD_2010_quintile","EIMD_2015_quintile","educfinh")
measset <- c("htval","wtval","bmival")
ageset <- c("agegad1", "agegad2", "agegch1","agegr1","age")
weightset <- c("wti_UKY1to11","astrata1","area")

indiv_data <- c(bpset,ethnset,saltset,medsset,hypset,incset,measset,ageset,weightset)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
sav11br <- sav11br[,..indiv_data]
sav4br <- sav4br[,..indiv_data]
sav6br <- sav6br[,..indiv_data]
sav8br <- sav8br[,..indiv_data]

val_labels(sav8br) <- NULL
val_labels(sav6br) <- NULL
val_labels(sav4br) <- NULL
val_labels(sav11br) <- NULL
#sav8br$seriali <- as.double(sav8br$seriali)
#sav8br$ethgrp5 <- as.double(sav8br$ethgrp5)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#this creates the whole table
ndns_1_11 <- rbindlist(list(sav4br,sav6br,sav8br,sav11br),fill = TRUE, use.names = TRUE)

```

```{r include=FALSE}
 #define factors for 1-4 
#saltset
ndns_1_11[, SaltChk := factor(SaltChk, levels = 1:8, labels = c("Salt",
	"Salt substitute",
	"Neither",
	"Item not applicable",
	"No answer/refused",
	"Don't know",
	"Qn not applicable to survey year",
	"Schedule not applicable"))]
 ndns_1_11[, SalHowC := factor(SalHowC, levels = 1:8,labels = c("Always",
"Usually",
"Sometimes",
	",Item not applicable",
	"No answer/refused",
	" Don't know",
"Qn not applicable to survey year",
"Schedule not applicable"))]
 ndns_1_11[, SltSHow := factor(SltSHow,levels = 1:8,labels = c("Always",
                                                            "Usually",
                                                            "Sometimes",
                                                            ",Item not applicable",
                                                            "No answer/refused",
                                                            " Don't know",
                                                            "Qn not applicable to survey year",
                                                            "Schedule not applicable"))]
 
 ndns_1_11[, vegetarn := factor(vegetarn, levels = 1:3, labels = c("vegetarian","vegan","not vegetarian"))]
```

```{r include=FALSE}
#medset
 ndns_1_11[, bpmedc2 := factor(bpmedc2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
 ndns_1_11[, bpmedd2 := factor(bpmedd2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, diur2 := factor(diur2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, beta2 := factor(beta2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, aceinh2 := factor(aceinh2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, calciumb2 := factor(calciumb2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
  ndns_1_11[, obpdrug2 := factor(obpdrug2,levels = 0:1,	labels = c("Not taking drug"
	,"Taking drug"))]
ndns_1_11[, PregNowB := factor(PregNowB,levels = 1:2,	labels = c("preg or breastfeeding"
	,"not preg or Bfeed"))]
 
 


 
 
 
#incset
  
 
ndns_1_11[,gor := factor(gor, levels = 1:12, labels = c("North East","North West","Yorkshire & The Humber","East Midlands","West Midlands","East of England","London","South East","South West","Wales","Scotland","Northern Ireland"))]


ndns_1_11[,region := factor(region, levels = 1:6, labels = c("England: North","England: Central/Midlands","England: South(including London)","Scotland","Wales","Northern Ireland"))] 
  
ndns_1_11[, EIMD_2007_quintile := factor(EIMD_2007_quintile, levels = 1:5)] 
ndns_1_11[,EIMD_2010_quintile := factor(EIMD_2010_quintile, levels = 1:5)] 
ndns_1_11[,EIMD_2015_quintile := factor(EIMD_2015_quintile, levels = 1:5)] 
ndns_1_11[,educfinh := factor(educfinh, levels = 1:8)] 
ndns_1_11[,nssec8 := factor(nssec8, levels = 1:8)] 
#bpset 
 
ndns_1_11[, "omsysval" := as.numeric(omsysval)]


 #hypeset
ndns_1_11[,hyper140_2 := factor(hyper140_2,levels = 1:9, labels = c(" Normotensive untreated",
" Normotensive treated",
	"Hypertensive treated",
"Hypertensive untreated",
"No answer/refused",
	"Don't know",
"Refused, attempted but not obtained, not attempted",
	 "Qn not applicable to survey year",
"Item not applicable")) ]
 ndns_1_11[, hibp140_2 := factor(hibp140_2, levels = 1:7, labels = c("Not high BP",
"High BP",
	"No answer/refused",
	"Don't know",
	"Refused, attempted but not obtained, not attempted",
	"Qn not applicable to survey year",
	"Item not applicable")) ]
ndns_1_11[, hyper1_2 := factor(hyper1_2,levels = 1:9, labels = c(" Normotensive untreated",
                                                          " Normotensive treated",
                                                          "Hypertensive treated",
                                                          "Hypertensive untreated",
                                                          "No answer/refused",
                                                          "Don't know",
                                                          "Refused, attempted but not obtained, not attempted",
                                                          "Qn not applicable to survey year",
                                                          "Item not applicable"))]
 ndns_1_11[, highbp1_2 := factor(highbp1_2, levels = 1:7, labels = c("Not high BP",
                                                              "High BP",
                                                              "No answer/refused",
                                                              "Don't know",
                                                              "Refused, attempted but not obtained, not attempted",
                                                              "Qn not applicable to survey year",
                                                              "Item not applicable")) ]
 
 
 
 
  
#ethnset  

ndns_1_11[ , ethgrp5 := factor(ethgrp5, levels = 1:5, labels = c( 'White'
                                                             , 'Mixed ethnic group'
                                                             , 'Black or Black British'
                                                             , 'Asian or asian British'
                                                             , 'Any other group'))]
ndns_1_11[ , ethgrp2 := factor(ethgrp2, levels = 1:2, labels = c( 'White'
                                                             , 'Non-white'))]




#ageset
ndns_1_11[,agegad1 := factor(agegad1, levels = 1:4 ,labels = c("16-24","25-49","50-64","65+ years"))]
ndns_1_11[,agegad2 := factor(agegad2, levels = 1:5, labels = c("16-18","19-34","35-49","50-64","65+ years"))]
ndns_1_11[,agegch1 := factor(agegch1, levels = 1:3, labels = c("8-10","11-12","13-15"))]
ndns_1_11[,agegr1 := factor(agegr1, levels = 1:5, labels = c("1.5-3 years","4-10 years","11-18 years","19-64 years","65+ years"))]
 




 
```



```{r include=FALSE}
#this identifies the variables wanted from the person level data sets
perslevdat <- c("seriali","Age","SurveyYear", "Sex","Country","Sodiummg","Calciummg","VitaminDµg","TotalEMJ", "Totalsugarsg","Glucoseg","Sucroseg","Fructoseg","Lactoseg","SOFTDRINKSLOWCALORIE","SOFTDRINKSNOTLOWCALORIE", "TEACOFFEEANDWATER")



```



```{r message=FALSE, warning=FALSE, include=FALSE}
#building the tables
sav4r <- sav4r[, ..perslevdat]
sav6r <- as.data.table(sav6r[, ..perslevdat])
sav8r <- as.data.table(sav8r[, ..perslevdat])
sav11r <- sav11r[,..perslevdat]


val_labels(sav8r) <- NULL
val_labels(sav6r) <- NULL
val_labels(sav4r) <- NULL
val_labels(sav11r) <- NULL

ndns_1_11r <- rbindlist(list(sav4r,sav6r,sav8r,sav11r),fill = TRUE, use.names = TRUE)

ndns_1_11 <- ndns_1_11[ ndns_1_11r, on = "seriali"]



```

```{r echo=FALSE}

ndns_1_11[, Sex := factor(Sex, levels = 1:2, labels = c("Male", "Female"))]
ndns_1_11[, Country := factor(Country)]



```



```{r load food diary data, message=FALSE, warning=FALSE, include=FALSE}
UPFNOVA <- data.table(read_excel("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/altquant/UPFNOVA.xlsx"))
#ndns_yr1_nutrientdatabank <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr1_nutrientdatabank.sav"))

#ndns_yr11_nutrient <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr11_nutrientdatabank_2021-03-19.sav"))

 ndns1_4a_foodleveldietarydata_uk_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_foodleveldietarydata_uk_v2.sav"))

ndns5_6a_foodleveldietarydata_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr5-6a_foodleveldietarydata_v2.sav"))
 
ndns7_8a_foodleveldietarydata_v2 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr7-8a_foodleveldietarydata.sav"))



 ndns_rp_yr9a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr9a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr10a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr10a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr11a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("/media/david/Elements1/contents/DOHmed/MPH/MPHModules/dissertation/alternativequant/altquant/UKDA-6533-spss/spss/spss25/ndns_rp_yr11a_foodleveldietarydata_uk_20210831.sav") )
```









```{r process food data into tables, message=FALSE, warning=FALSE, include=FALSE}
#taking data from diet diary

UPFNOVA[,`Nova group`:= factor(`Nova group`,levels = 1:4)]
UPFNOVA$SubFoodGroupCode <- UPFNOVA$`food group number`

#specifically coded food groups

#combine with key for Nova Group
#first 1-4
foodlevedat4s <- ndns1_4a_foodleveldietarydata_uk_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
#foodlevedat4s[ ,seriali  := factor(seriali)]
foodlevedat4s<- foodlevedat4s[UPFNOVA, on= "SubFoodGroupCode"]
#summary(foodlevedat4s)

foodlevedat4s<- foodlevedat4s[,c("seriali","TotalGrams","Nova group","EnergykJ")]

#years 5-6
foodlevedat6s <- ndns5_6a_foodleveldietarydata_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat8s <- ndns7_8a_foodleveldietarydata_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat6s<- foodlevedat6s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat8s<- foodlevedat8s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat6s<- foodlevedat6s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat8s<- foodlevedat8s[,c("seriali","TotalGrams","Nova group","EnergykJ")]


#finally years 9, 10, 11
foodlevedat9s <- ndns_rp_yr9a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat10s <- ndns_rp_yr10a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]
foodlevedat11s <- ndns_rp_yr11a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams","EnergykJ")]

#combine with key for Nova Group

foodlevedat9s<- foodlevedat9s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat10s<- foodlevedat10s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat11s<- foodlevedat11s[UPFNOVA, on= "SubFoodGroupCode"]



foodlevedat9s<- foodlevedat9s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat10s<- foodlevedat10s[,c("seriali","TotalGrams","Nova group","EnergykJ")]
foodlevedat11s<- foodlevedat11s[,c("seriali","TotalGrams","Nova group","EnergykJ")]

```

```{r message=FALSE, warning=FALSE, include=FALSE}
#heatmap(foodlevedat9s, Colv = "TotalGrams")

```



```{r function defn, message=FALSE, warning=FALSE, include=FALSE}
#To work out the gram weight amount of food intake by each individual, first the diary entries for each individual are totalled up.
#The total gram weight value of intake of each food is then worked out as a percentage of the total intake.



## First I write the function to process the individual entries. This totals the intake in grams and identifies the proportion by nova type.



IndividDiet <- function(DT,individ) {
  pers <- DT[seriali==individ]
pers
oneweight<-pers[, round(sum(TotalGrams),2), by=`Nova group`][order(`Nova group`)]

setnames( oneweight, 'V1','Totalg')
 oneweight[,pcnt := round((Totalg/sum(Totalg)*100),2)]

  return(na.exclude(oneweight))
}


```



```{r year 9 process, message=FALSE, warning=FALSE, include=FALSE}
#All these individual calculations are then built back up into tables.
#This is done for years 9-11 and then 1-4.

### This section uses the function designed above to process the table
name <-foodlevedat9s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat9s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat9sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat9sp



```

```{r year 10 process, warning=FALSE, include=FALSE}

name <-foodlevedat10s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat10s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat10sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat10sp


```

```{r year 11 process, message=FALSE, warning=FALSE, include=FALSE}

name <-foodlevedat11s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat11s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat11sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat11sp

#then combining the outputs 9-11
foodleveldat910sp <- rbind(foodleveldat9sp,foodleveldat10sp)
foodleveldat911sp <- rbind(foodleveldat11sp, foodleveldat910sp)
```

```{r year 5-6 process, warning=FALSE, include=FALSE}

name <-foodlevedat6s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat6s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat6sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat6sp


```

```{r year 7-8 process, warning=FALSE, include=FALSE}

name <-foodlevedat8s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat8s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat8sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat8sp


```

```{r slow processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat4s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat4s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat4sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
#foodleveldat4sp



```



```{r energy function defn, message=FALSE, warning=FALSE, include=FALSE}
#The process can be done for food level energy intake also.

## First I write the function to process the individual entries. This totals the intake in grams and identifies the proportion by nova type.



IndividDietE <- function(DT,individ) {
  pers <- DT[seriali==individ]
pers
oneweight<-pers[, round(sum(EnergykJ),2), by=`Nova group`][order(`Nova group`)]

setnames( oneweight, 'V1','EnergykJ')
 oneweight[,Epcnt := round((EnergykJ/sum(EnergykJ)*100),2)]

  return(na.exclude(oneweight))
}


```

```{r Eyear 9 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat9s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat9s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat9spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat9spE



```

```{r Eyear 10 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat10s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat10s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat10spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat10spE



```

```{r Eyear 11 process, message=FALSE, warning=FALSE, include=FALSE}
### This section uses the function designed above to process the table
name <-foodlevedat11s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat11s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat11spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
foodleveldat11spE


#then combining the outputs 9-11
foodleveldat910spE <- rbind(foodleveldat9spE,foodleveldat10spE)
foodleveldat911spE <- rbind(foodleveldat11spE, foodleveldat910spE)
```

```{r Eslow processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat4s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat4s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat4spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat4spE



```

```{r E8 processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat8s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat8s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat8spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat4spE



```

```{r E6 processing four years, echo=FALSE, warning=FALSE}

name <-foodlevedat6s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDietE(foodlevedat6s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat6spE <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('EnergykJ',"Epcnt"))
#foodleveldat6spE



```


```{r adding to earlier data2, echo=FALSE, message=FALSE, warning=FALSE}

# combine with data from version10a.rmd

#sav11rpb <- readRDS("sav11rp.rds")


setkey(foodleveldat911sp, seriali)
setkey(foodleveldat911spE, seriali)

foodleveldat911spE <-foodleveldat911spE[foodleveldat911sp, by= seriali]
#sav11rp <-sav11rp[foodleveldat911spE, by= seriali]
#sav11rp$ED <- sav11rp[,.("Epcnt_4"/"pcnt_4")]

#sav4rp <- readRDS("sav4rp.rds")

#setkey(sav4rp,seriali)
setkey(foodleveldat4sp,seriali)
setkey(foodleveldat4spE,seriali)
foodleveldat4spE<-foodleveldat4spE[foodleveldat4sp, by= seriali]
#sav4rp <-sav4rp[foodleveldat4spE, by= seriali]


#setkey(sav6rp,seriali)
setkey(foodleveldat6sp,seriali)
setkey(foodleveldat6spE,seriali)
foodleveldat6spE <-foodleveldat6spE[foodleveldat6sp, by= seriali]
#sav6rp <-sav6rp[foodleveldat6spE, by= seriali]



#setkey(sav8rp,seriali)
setkey(foodleveldat8sp,seriali)
setkey(foodleveldat8spE,seriali)
foodleveldat8spE <-foodleveldat8spE[foodleveldat8sp, by= seriali]
#sav8rp <-sav8rp[foodleveldat8spE, by= seriali]

ndns_1_11food <- rbindlist(list(foodleveldat8spE,foodleveldat4spE,foodleveldat6spE,foodleveldat911spE))
```



```{r adding to earlier data, echo=FALSE, message=FALSE, warning=FALSE}
# combine with data from version10a.rmd
setkey(ndns_1_11food, seriali)
setkey(ndns_1_11, seriali)
ndns_1_11 <- ndns_1_11[ndns_1_11food, by= seriali]
```



```{r message=FALSE, warning=FALSE, include=FALSE}
#data(ndns_1_11)


ndns_1_11d <- svydesign(ids = ~area, weights = ~ wti_UKY1to11, strata=~astrata1, data = ndns_1_11, nest = TRUE)

#summary(ndns_1_11d)
#sysMean <- svymean(~omsysval, ndns_1_11d, na.rm = TRUE)
#sysMean
#Agemean <- svymean(~Age, ndns_1_11d, na.rm = TRUE)
#Agemean
#sexmean <- svymean(~Sex, ndns_1_11d, na.rm = TRUE)
#sexmean
#Namean <-  svymean(~Sodiummg, ndns_1_11d, na.rm = TRUE)
#Namean
#Epmean <- svymean(~Epcnt_4, ndns_1_11d, na.rm = TRUE)
#Epmean


```




```{r echo=FALSE}
#Exclusions


#ndns_1_11 <-ndns_1_11[(hyper1_2 == " Normotensive untreated" ), ]
ndns_1_11e <- ndns_1_11[omsysval <= 150]
#ndns_1_11hypo <- subset(ndns_1_11, ndns_1_11$omsysval <= 150)

#there are no pregnant or breast feeding participants
#ndns_1_11n <- ndns_1_11n[(PregNowB != "preg or breastfeeding" ), ]


#medication affecting bp
ndns_1_11e <- ndns_1_11e[diur2 == "Not taking drug"]
ndns_1_11e <-ndns_1_11e[beta2 == "Not taking drug"]
ndns_1_11e <-ndns_1_11e[aceinh2 == "Not taking drug"]
ndns_1_11e <-ndns_1_11e[calciumb2 =="Not taking drug"]
ndns_1_11e <- ndns_1_11e[obpdrug2 == "Not taking drug"]

#ndns_1_11 <- ndns_1_11[Country == "England"]
ndns_1_11e <- ndns_1_11e[Age >18]
ndns_1_11e <- svydesign(ids = ~area, weights = ~ wti_UKY1to11, strata=~astrata1, data = ndns_1_11, nest = TRUE)
```
#  Results Section
##  Results Introduction

The results derive from the method outlined above and follow the pattern described. 
I will try to discuss the results having already described the method.



## Descriptive Data Analysis

This first table highlights the data from the years 2008-2012 at the beginning of the data collection.
```{r echo=FALSE, message=FALSE, warning=FALSE}
first4 <- subset(ndns_1_11d,SurveyYear <= 4 )
Na4Mean <- svymean(~Sodiummg, first4, na.rm = TRUE)
Na4Mean
Sys4mean <- svymean(~omsysval, first4, na.rm = TRUE)
Sys4mean
Ep4mean <- svymean(~Epcnt_4, first4, na.rm = TRUE)
Ep4mean
#svytable(~Sodiummg+Epcnt_4+omsysval, first4, Ntotal =NULL, round = FALSE)

first4_d <- as_survey(first4)
class(first4_d)
subset(ndns_1_11,SurveyYear <=4 ) %>% tbl_summary(include = c(Sodiummg,Epcnt_4,pcnt_4,omsysval),
                 statistic = list(all_continuous() ~"{mean} ({sd})"))
first4_d %>% tbl_svysummary(include = c(Sodiummg,Epcnt_4,pcnt_4,omsysval),
                 statistic = list(all_continuous() ~"{mean} ({sd})"))

```
The second table show the data from the second cohort 2017-19.

```{r echo=FALSE, message=FALSE, warning=FALSE}
last3 <- subset(ndns_1_11d,SurveyYear >= 9 )
Na9Mean <- svymean(~Sodiummg, last3, na.rm = TRUE)
Na9Mean
sys9mean <- svymean(~omsysval, last3, na.rm = TRUE)
sys9mean
Ep9mean <- svymean(~Epcnt_4, last3, na.rm = TRUE)
Ep9mean

last3_d <- as_survey(last3)
class(last3_d)

subset(ndns_1_11,SurveyYear >= 9 ) %>% tbl_summary(include = c(Sodiummg,Epcnt_4,omsysval),
                 statistic = list(all_continuous() ~"{mean} ({sd})"))
last3_d %>% tbl_svysummary(include = c(Sodiummg,Epcnt_4,omsysval),
                 statistic = list(all_continuous() ~"{mean} ({sd})"))
```
These tables show that there has been a change between the two cohorts.
They also show that the weighted samples are different to raw samples identifying the importance of analysing the data across weighted groups.




```{r echo=FALSE, message=FALSE, warning=FALSE}
#bpNa <- data.table("2010"=c(ndns_1_11d[SurveyYear <= 4,Sodiummg*wti_UKY1to11]),"2019" =c(ndns_1_11d[SurveyYear >= 9,Sodiummg*wti_UKY1to11]))
#boxplot(bpNa, main = "Sodiummg 2010 vs 2019", xlab = "year", ylab = "Sodium mg")
#bpgram <-data.table("2010" =c(ndns_1_11d[SurveyYear <= 4,pcnt_4*wti_UKY1to11]),"2019" = c(ndns_1_11d[SurveyYear >= 9,pcnt_4*wti_UKY1to11]))
#boxplot(bpgram, main = "Weight Percent UPF 2010 vs 2019", xlab = "year", ylab = "pcnt_4")
#bpE <-data.table("2010" =c(ndns_1_11d[SurveyYear <= 4,Epcnt_4*wti_UKY1to11/4]),"2019" = c(ndns_1_11d[SurveyYear >= 9,Epcnt_4*wti_UKY1to11/4]))
#boxplot(bpE, main = " Energy Percent UPF 2010 vs 2019", xlab = "year", ylab = "Epcnt_4")


#bpOS <- data.table("2010" = c(ndns_1_11[SurveyYear <=4,omsysval*wti_UKY1to11]),"2019"=c(ndns_1_11[SurveyYear >=9,omsysval*wti_UKY1to11]))
#boxplot(bpOS,main = "omsysval 2010 vs 2019",xlab = "year", ylab = "omsysval")
#bpOD <-data.table("2010" = c(ndns_1_11[SurveyYear <=4,omdiaval*wti_UKY1to11]),"2019" = c(ndns_1_11[SurveyYear >=9,omdiaval*wti_UKY1to11]))
#boxplot(bpOD, main = "omdiaval 2010 vs 2019", xlab = "year", ylab = "omdiaval")


```

```{r echo=FALSE, message=FALSE, warning=FALSE}

#keydata

#svyboxplot(Epcnt_4~SurveyYear,ndns_1_11d)
#svyboxplot(Sodiummg~SurveyYear,ndns_1_11d)
#svyboxplot(omsysval~SurveyYear,ndns_1_11d, na.rm = TRUE)
```
These box plots show how the percentage of energy derived from UPF, the sodium intake, and the Systolic bp have changed over the years.
The graphs show that there is not a visible difference between the years. Statistical analysis will follow.


## Comparative analysis

### Comparison of key variables

####  comparing UPF and Sodium intake calculated from diet

In order to confirm there has been a change in intake a t.test compares the means of the two samples.
One compares the means of sodium in years 1-4 with sodium in years 9-11.

The second compares the means of pcnt UPF intake in over the same periods.
A third compares the percentage energy provided by UPF.

```{r echo=FALSE}

ttestNa <- t.test(ndns_1_11[SurveyYear >= 9,Sodiummg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Sodiummg*wti_UKY1to11])
ttestpcnt_4 <-t.test(ndns_1_11[SurveyYear >= 9,pcnt_4*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,pcnt_4*wti_UKY1to11])
ttestEpcnt_4 <-t.test(ndns_1_11[SurveyYear >= 9,Epcnt_4*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Epcnt_4*wti_UKY1to11])


data.table("Var" = c("Epcnt_4","pcnt_4","Na"),"statistic" = signif(c(ttestEpcnt_4$statistic,ttestpcnt_4$statistic,ttestNa$statistic), digits = 4), "p.value" = signif(c(ttestEpcnt_4$p.value,ttestpcnt_4$p.value,ttestNa$p.value), digits = 4))
ttestpcnt_4
ttestEpcnt_4
ttestNa
```

It seems the mean percentage UPF intake changes from 48.8% to 59.2% energy and this increase is statistically significant.
The mean sodium intake has changed from  2156.30 mg to 2574.33 mg and is also statistically significant with a p value less than 0.05.

#### what about outcome BP?

The next t tests compare mean systolic values in the two time periods and then the mean diastolic values.

```{r echo=FALSE}



ttestOS <- t.test(ndns_1_11[SurveyYear >=9,omsysval*wti_UKY1to11], ndns_1_11[SurveyYear <=4,omsysval*wti_UKY1to11])
ttestOD <- t.test(ndns_1_11[SurveyYear >=9,omdiaval*wti_UKY1to11], ndns_1_11[SurveyYear <=4,omdiaval*wti_UKY1to11])
data.table("Var" = c("Sys","Dia"),"statistic" = signif(c(ttestOS$statistic,ttestOD$statistic), digits= 4), "p.value" = signif(c(ttestOS$p.value,ttestOD$p.value), digits = 4))

ttestOS
```

There is a change in mean systolic from 122-152 mmHg with a p value of 3.112e -7.

In summary there is statistically significant change in UPF and Na intake and also in both systolic and diastolic pressures.

Has another factor affected the BP change ?

#### Statistical analysis of other variables

How are variables distributed between the two cohorts. 
The NDNS dataset was weighted to keep many of these the same between datasets.
Continuous variables are assessed using ttests and categorical variables using chi squared tests to give p.values.

######  Age and Sex
There is 

```{r echo=FALSE}
#from pers data
#continuous data
#ttestage <-svyttest(~first4$Age, ~last9$Age, )
ttestage <-t.test(ndns_1_11[SurveyYear >= 9,Age*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Age*wti_UKY1to11])
ttestca <- t.test(ndns_1_11[SurveyYear >= 9,Calciummg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Calciummg*wti_UKY1to11])
ttestts <-t.test(ndns_1_11[SurveyYear >= 9,Totalsugarsg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Totalsugarsg*wti_UKY1to11])
ttestgl <-t.test(ndns_1_11[SurveyYear >= 9,Glucoseg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Glucoseg*wti_UKY1to11])
ttestfr <- t.test(ndns_1_11[SurveyYear >= 9,Fructoseg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Fructoseg*wti_UKY1to11])
ttestsu <-t.test(ndns_1_11[SurveyYear >= 9,Sucroseg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Sucroseg*wti_UKY1to11])
ttestla <-t.test(ndns_1_11[SurveyYear >= 9,Lactoseg*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,Lactoseg*wti_UKY1to11])
ttestlo <-t.test(ndns_1_11[SurveyYear >= 9,SOFTDRINKSLOWCALORIE*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,SOFTDRINKSLOWCALORIE*wti_UKY1to11])
ttestnlo <-t.test(ndns_1_11[SurveyYear >= 9,SOFTDRINKSNOTLOWCALORIE*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,SOFTDRINKSNOTLOWCALORIE*wti_UKY1to11])
ttesttc<- t.test(ndns_1_11[SurveyYear >= 9,TEACOFFEEANDWATER*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,TEACOFFEEANDWATER*wti_UKY1to11])

confounders <- data.table("name" = c("Age","Calciummg","Totalsugarsg","Glucoseg","Fructoseg","Sucroseg", "Lactoseg","SOFTDRINKSLOWCALORIE","SOFTDRINKSNOTLOWCALORIE","TEACOFFEEANDWATER"), 
 "pvalue" = signif(c(ttestage$p.value, ttestca$p.value,ttestts$p.value,ttestgl$p.value,ttestfr$p.value,ttestsu$p.value,ttestla$p.value,ttestlo$p.value,ttestnlo$p.value,ttesttc$p.value), digits = 4), "statistic" = signif(c(ttestage$statistic, ttestca$statistic,ttestts$statistic,ttestgl$statistic,ttestfr$statistic,ttestsu$statistic,ttestla$statistic,ttestlo$statistic,ttestnlo$statistic,ttesttc$statistic), digits = 4))
confounders

```

The age of the two datasets has changed but not in a statistically significant way.
Intake seems to be significantly different between the datasets.

There has been a change in the intake of soft drinks,not low calorie.

```{r echo=FALSE, warning=FALSE}

#categorical data
stable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,Sex])
stable4 <- xtabs(~ ndns_1_11[SurveyYear <= 4,Sex])
stabler <- chisq.test(rbind(stable11 , stable4))


#ctable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,Country*wti_UKY1to11])
#ctable4 <- xtabs(~ ndns_1_11[SurveyYear <= 4,Country*wti_UKY1to11])
#ctabler <- chisq.test(rbind(ctable4,ctable11))

catconfounders <- data.table("name" = c("Sex"),"pvalue" = signif(c(stabler$p.value),digits = 4))

catconfounders


plot(ndns_1_11[SurveyYear >= 9,Sex], main = "2019", caption = "Bar Plot of Sex in 2019 cohort")
plot( ndns_1_11[SurveyYear <= 4,Sex], main = "2010",caption = "Bar Plot of Sex in 2019 cohort")
#boxplot(ndns_1_11[,Sex], ndns_1_11[,SurveyYear])
#plot(sav11rp[,Country], main = "2019")
#plot(sav4rp[,Country], main = "2010")

```

There is a statistically significant change in the sex distribution of the two groups.
This might be due to differences in the numbers of excluded participants.
In particular there may be more younger people and women taking e.g. bblockers in one group.

```{r echo=FALSE}
#ttests for continuous data
ttestht <- t.test(ndns_1_11[SurveyYear >= 9,htval*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,htval*wti_UKY1to11])
ttestwt <- t.test(ndns_1_11[SurveyYear >= 9,wtval*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,wtval*wti_UKY1to11])
ttestbmi <- t.test(ndns_1_11[SurveyYear >= 9,bmival*wti_UKY1to11], ndns_1_11[SurveyYear <= 4,bmival*wti_UKY1to11])

confounders2 <- data.table("name" = c("htval","wtval","bmival"),"pvalue" = signif(c(ttestht$p.value,ttestwt$p.value,ttestbmi$p.value),digits = 4),"statistic" = signif(c(ttestht$statistic,ttestwt$statistic,ttestbmi$statistic),digits = 4))
confounders2




```

This table suggests that there is a significant difference between the height, and bmi of the groups.

```{r echo=FALSE, warning=FALSE}

#chisquared for categorical data
#saltset
vtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,vegetarn])
vtable4 <- xtabs(~ ndns_1_11[SurveyYear <= 4,vegetarn])
vtabler <- chisq.test(rbind(vtable11 , vtable4))

#cktable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,SaltChk*wti_UKY1to11])
#cktable4 <- xtabs(~ ndns_1_11[SurveyYear <= 4,SaltChk*wti_UKY1to11])
#cktabler <- chisq.test(rbind(cktable11 , cktable4))

#hwtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,SaltHowC*wti_UKY1to11])
#hwtable4 <- xtabs(~ ndns_1_11[SurveyYear <= 4,SaltHowC*wti_UKY1to11])
#hwtabler <-chisq.test(rbind(hwtable11 , hwtable4))

#swtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,SltSHow*wti_UKY1to11])
#swtable4 <- xtabs(~nndns_1_11[SurveyYear <= 4,SltSHow*wti_UKY1to11])
#swtabler <- chisq.test(rbind(swtable11 , swtable4))

#catconfounders2 <- data.table("name" = c("vegetarn","SaltChk","SalHowC","SltShow"),"p.value" = signif(c(vtabler$p.value,cktabler$p.value,hwtabler$p.value,swtabler$p.value ),digits = 4))
catconfounders2 <- data.table("name" = c("vegetarn"),"p.value" = signif(c(vtabler$p.value ),digits = 4),"statistic" = signif(c(vtabler$statistic ),digits = 4))
catconfounders2

```

These values identify a significant difference in the number of vegetarians

```{r echo=FALSE, warning=FALSE}

#ethnset
e5table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,ethgrp5])
e5table4 <- xtabs(~ndns_1_11[SurveyYear <=4,ethgrp5])
e5tabler <- chisq.test(rbind(e5table11 , e5table4))

e2table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,ethgrp2])
e2table4 <- xtabs(~ndns_1_11[SurveyYear <=4,ethgrp5])
e2tabler <- chisq.test(rbind(e2table11 , e2table4))

#incset

#n8table11 <- xtabs(~sav11rp[,nssec8])
#n8table4 <- xtabs(~sav4rp[, nssec8])
#n8tabler<- chisq.test(rbind(n8table11 , n8table4))

#gtable11 <- xtabs(~sav11rp[,GOR])
#gtable4 <- xtabs(~sav4rp[,gor])
#gtabler <- chisq.test(rbind(gtable11 , gtable4))

#rtable11 <- xtabs(~sav11rp[,region])
#rtable4 <- xtabs(~sav4rp[, region])
#rtabler <-chisq.test(rbind(rtable11 , rtable4))


q07table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,EIMD_2007_quintile])
q07table4 <- xtabs(~ndns_1_11[SurveyYear<=4,EIMD_2007_quintile])
q07tabler <-chisq.test(rbind(q07table11 , q07table4))

q10table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,EIMD_2010_quintile])
q10table4 <- xtabs(~ndns_1_11[SurveyYear <=4,EIMD_2007_quintile])
q10tabler <-chisq.test(rbind(q10table11 , q10table4))

q15table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,EIMD_2007_quintile])
q15table4 <- xtabs(~ndns_1_11[SurveyYear <=4,EIMD_2007_quintile])
q15tabler <-chisq.test(rbind(q15table11 , q15table4))

edutable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,educfinh])
edutable4 <- xtabs(~ndns_1_11[SurveyYear <=4,educfinh])
edutabler <-chisq.test(rbind(edutable11 , edutable4))


#catconfounders3a <- data.table("name" = c("GOR","region"),"p.value" =signif(c( gtabler$p.value, rtabler$p.value ),digits = 4))

catconfounders3b <- data.table("name" = c("ethgrp5","ethgrp2"),"statistic" =signif(c(e5tabler$statistic, e2tabler$statistic),digits = 4),"p.value" =signif(c(e5tabler$p.value, e2tabler$p.value),digits = 4))

catconfounders3c <- data.table("name" = c("EIMD_2007_quintile","EIMD_2010_quintile","EIMD_2015_quintile"),"statistic" =signif(c(q07tabler$statistic, q10tabler$statistic, q15tabler$statistic),digits = 4),"p.value" =signif(c(q07tabler$p.value, q10tabler$p.value, q15tabler$p.value),digits = 4))

catconfounders3d <- data.table("name" = c("educfin"),"statistic" =signif(c(edutabler$statistic),digits = 4),"p.value" =signif(c(edutabler$p.value),digits = 4))

#catconfounders3a
catconfounders3b
catconfounders3c
catconfounders3d
```

There are differences in ethnicity as divided into 5 subgroups.
The differences in qimd, are not statistically significant.
There is a difference in the age of finishing education.

```{r echo=FALSE, warning=FALSE}

#hypeset

h1table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,hyper1_2])
h1table4 <- xtabs(~ndns_1_11[SurveyYear <=4,hyper1_2])
h1tabler <- chisq.test(rbind(h1table11 , h1table4))

h14table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,hyper140_2])
h14table4 <- xtabs(~ndns_1_11[SurveyYear <=4,hyper140_2])
h14tabler <- chisq.test(rbind(h14table11 , h14table4))

#hi1table11 <- xtabs(~sav11rp[,highbp1_2])
#hi1table4 <- xtabs(~sav4rp[, highbp1 ])
#hi1tabler <- chisq.test(rbind(hi1table11 , hi1table4))

hi14table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,hibp140_2])
hi14table4 <- xtabs(~ndns_1_11[SurveyYear <=4,hibp140_2])
hi14tabler <- chisq.test(rbind(hi14table11 , hi14table4))

#ageset

ad1table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,agegad1])
ad1table4 <- xtabs(~ndns_1_11[SurveyYear <=4,agegad1])
ad1tabler <- chisq.test(rbind(ad1table11 , ad1table4))

ad2table11 <- xtabs(~ndns_1_11[SurveyYear >= 9,agegad2])
ad2table4 <- xtabs(~ndns_1_11[SurveyYear <=4,agegad2])
ad2tabler <- chisq.test(rbind(ad2table11 , ad2table4))

#ch1table11 <- xtabs(~sav11rp[,agegch1])
#ch1table4 <- xtabs(~sav4rp[, agegch1 ])
#ch1tabler <-chisq.test(rbind(ch1table11 , ch1table4))

#ag1table11 <- xtabs(~sav11rp[, agegr1])
#ag1table4 <- xtabs(~sav4rp[, agegr1])
#ag1tabler <- chisq.test(rbind(ag1table11 , ag1table4))

catconfounders4a <- data.table("name" = c("agegad1","agegad2"),"p.value" = signif( c( ad1tabler$p.value , ad2tabler$p.value), digits = 4))

catconfounders4b <- data.table("name" = c("hyper1","hyper140","highbp1","hibp140"),"p.value" = signif( c(h1tabler$p.value, h14tabler$p.value, 0, hi14tabler$p.value), digits = 4))


catconfounders4a
#catconfounders4b
```

The age groups show some discrepancy with the p value significant only in the child age groups.

```{r echo=FALSE, warning=FALSE}

#medset

mctable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,bpmedc2])
mctable4 <- xtabs(~ndns_1_11[SurveyYear <=4,bpmedc2 ])
mctabler <- chisq.test(rbind(mctable11 , mctable4))

mdtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9, bpmedd2])
mdtable4 <- xtabs(~ndns_1_11[SurveyYear <=4, bpmedd2])
mdtabler <- chisq.test(rbind(mdtable11 , mdtable4))

ditable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,diur2])
ditable4 <- xtabs(~ndns_1_11[SurveyYear <=4, diur2])
ditabler <- chisq.test(rbind(ditable11 , ditable4))

betable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,beta2])
betable4 <- xtabs(~ndns_1_11[SurveyYear <=4, beta2])
betabler <- chisq.test(rbind(betable11 , betable4))

cbtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,calciumb2])
cbtable4 <- xtabs(~ndns_1_11[SurveyYear <=4, calciumb2])
cbtabler <-chisq.test(rbind(cbtable11 , cbtable4))

acetable11 <- xtabs(~ndns_1_11[SurveyYear >= 9, aceinh2])
acetable4 <- xtabs(~ndns_1_11[SurveyYear <=4, aceinh2])
acetabler <- chisq.test(rbind(acetable11 , acetable4))

othtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,obpdrug2])
othtable4 <- xtabs(~ndns_1_11[SurveyYear <=4, obpdrug2])
othtabler <- chisq.test(rbind(othtable11 , othtable4))

prtable11 <- xtabs(~ndns_1_11[SurveyYear >= 9,PregNowB])
prtable4 <- xtabs(~ndns_1_11[SurveyYear <=4, PregNowB])
prtabler <- chisq.test(rbind(prtable11 , prtable4))

catconfounders5 <- data.table("name" = c("bpmedc","bpmedd","diur","beta","calciumb","aceinh","obpdrug","PregNowB"), "p.value" = signif(c(mctabler$p.value, mdtabler$p.value, ditabler$p.value, betabler$p.value, cbtabler$p.value, acetabler$p.value, othtabler$p.value, prtabler$p.value ),digits = 4))

#catconfounders5

```

## Regression Analysis

### linear regression

Simple linear regression equations look for the relationship between the dependant variable, and the independent variable.
For these I am looking at the whole dataset 

```{r echo=FALSE}
#svyplot(omsysval~Sodiummg, ndns_1_11e )

#svyplot(omsysval~EnergykJ_4, ndns_1_11e)
#svyplot(omsysval~pcnt_4, ndns_1_11e)

```

The regression models are examined for Sodium and UP against BP.
These use the populations where participants have been excluded. 
(analysis including these makes no difference!!)



 omsysval is compared to EnergykJ 

```{r echo=FALSE}
#linear regression models
#UPF Energy % only
lm1E <- svyglm(omsysval ~ EnergykJ_4, ndns_1_11e)



lm1E

#anova(lm1E)
#AIC(lm1E)


#sensemakr(lm1E, treatment = "EnergykJ_4")

```

then sodiummg

```{r echo=FALSE}
#linear regression models
#sodium only
lm1na <- svyglm(omsysval ~ Sodiummg, ndns_1_11e)





(lm1na)

anova(lm1na)
AIC(lm1na)


#sensemakr(lm1na, treatment = "Sodiummg")


```

There are relationships between Na and g pcnt as well as E pcnt and omsysval .


In conclusion the linear regression models show that there are statistically significant positive correlations between the systolic BP and each of the key variables.



### multi variable regression

This uses a model of variables.
It can highlight the contributions of each variable.
The intention is to develop an optimal model which mathematically describes the situation.

In particular the research question asks about the relationship between Sodium and UPF intake with BP.
The models will reflect this question with models looking to include or exclude particular variables.
Comparisons between these models are then made using sensitivity analysis, identifying how sensitive the model is to sodium, or other factors

This first model looks at the relationships between BP and Age and Sex education and IMD all of which may have an effect on BP. This model excludes UPF and Na.

```{r echo=FALSE}
#linear regression models
#sodium only
lm1AS <- svyglm(omsysval ~ Age + Sex +bmival +educfinh + EIMD_2010_quintile , ndns_1_11e, na.action = na.exclude)





(lm1AS)

anova(lm1AS)
#AIC(lm1AS)
lm1AS$aic
#sensemakr(lm1AS, treatment = "Age")

#sensemakr(lm1AS, treatment = "Sex")

```
This first model shows that all these variables, Age, Sex, education, IMD,  and bmi, give statistically significant coefficients for the model which suggests that they do have an important part to play in any optimal model.

The next model adds Sodiummg.

```{r echo=FALSE}
lmM <- svyglm(omsysval ~ Age +Sex +Sodiummg + bmival+ educfinh+ EIMD_2010_quintile , ndns_1_11e,na.action = na.exclude)

lmM


anova(lmM)
#AIC(lmM)


#sensemakr(lmM, treatment = "Age")


```
This second model gives Sodiummg, educfinh, and IMD statistical significance. VitaminD shows no statistical significance, TotalEMJ and sqrt(pcnt) and ethgrp2 all have limited significance.



Now we add UPF as total energy from nova 4 or UPF

```{r echo=FALSE, message=FALSE, warning=FALSE}
lmM2 <- svyglm(omsysval ~ Age +Sex +Sodiummg + EnergykJ_4+ bmival  +educfinh+ EIMD_2010_quintile , ndns_1_11e,na.action = na.exclude)



lmM2

anova(lmM2)
#AIC(lmM2)

```

UPF does not seem significant...

but when removing sodiummg

```{r echo=FALSE, warning=FALSE}
lmMna <- svyglm(omsysval ~ Age +Sex + EnergykJ_4+ bmival  +educfinh+ EIMD_2010_quintile  , ndns_1_11e,na.action =na.exclude)



lmMna


Am1 <-anova(lmMna)
#AIC(lmMna)
Am1
```
the UPF becomes significant! 
This suggests that the effect of UPF is mediated by Sodium!!

comparing AIC for tghese three models 

```{r}
lmM$aic
lmMna$aic
lmM2$aic

#AIC(lmM,lmMna,lmM2)


```
we find that the lowest AIC is given by the model without UPF!!
Though all the models with UPF have a lower aic than the model without.



## Summary of Results

There is a table with summary values for the key variables across the dataset.

Statistical analysis of the key variables shows the change in all the variables between the two time periods.

Confounding variables are analysed and show if there has been a significant change in the balance of the populations.


Regression shows a degree of association between the BP and UPF intake by weight and by energy.
It also shows the same for sodium intake.

Using Anova analysis of different multi variable regression models the key variables are significant for sodium in several models, and sometimes for UPF.


## Conclusion

The percentage by weight of NOVA group 4 foods increased from 2008 to 2019.
The percentage by energy of NOVA group 4 foods increased from 2008 to 2019.
The mean sodium intake in mg increased between the two time periods.
The systolic and diastolic BP have increased between the two time periods.

In each period there is a correlation between systolic BP and sodium intake.
In each period there is a correlation between systolic BP and UPF intake.

The regression models identify that age and sex are statistically significant contributors to the BP.

The regression models identify that sodium intake is an important contributor to any optimal model. That when sodium intake is present UPF intake is no longer significant, but still has some effect. 

