---
title: "upf"
author: "doh"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## UPF

```{r cars}
library(data.table)
library(readxl)
library(haven)
```

## upf database

```{r pressure}
UPFNOVA <- data.table(read_excel("altquant/UPFNOVA.xlsx"))
#ndns_yr1_nutrientdatabank <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr1_nutrientdatabank.sav"))

#ndns_yr11_nutrient <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_yr11_nutrientdatabank_2021-03-19.sav"))

 ndns1_4a_foodleveldietarydata_uk_v2 <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_rp_yr1-4a_foodleveldietarydata_uk_v2.sav"))

 ndns_rp_yr9a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_rp_yr9a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr10a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_rp_yr10a_foodleveldietarydata_uk_20210831.sav"))
 ndns_rp_yr11a_foodleveldietarydata_uk_20210831 <- data.table(read_sav("UKDA-6533-spss/spss/spss25/ndns_rp_yr11a_foodleveldietarydata_uk_20210831.sav") )
```

```{r}
 UPFNOVA

#labels = c("unprocessed","ingredients","processed","ultraprocessed","complex","supplements")
UPFNOVA[,`Nova group`:= factor(`Nova group`,levels = 1:4)]
UPFNOVA$SubFoodGroupCode <- UPFNOVA$`food group number`
plot(UPFNOVA$`Nova group`)
```

```{r}

#taking data from diet diary
#specifically coded food gorups

foodlevedat4s <- ndns1_4a_foodleveldietarydata_uk_v2[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams")]
summary(foodlevedat4s$seriali)
#combine with key for Nova Group
foodlevedat4s[ ,seriali  := factor(seriali)]
foodlevedat4s<- foodlevedat4s[UPFNOVA, on= "SubFoodGroupCode"]
summary(foodlevedat4s)

foodlevedat4s<- foodlevedat4s[,c("seriali","TotalGrams","Nova group")]
#identify total intake in grams by nova group
weights <-foodlevedat4s[, sum(TotalGrams), by=`Nova group`]


foodlevedat4s[plot(`Nova group`)]
plot(weights, ylab = "Grams")
#foodlevedat4s[plot(`Nova group`, TotalGrams)]
```

So there is a significant amount of UPF use. Has this has changed ?

```{r}

#taking data from diet diary
#specifically coded food gorups

foodlevedat9s <- ndns_rp_yr9a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams")]
foodlevedat10s <- ndns_rp_yr10a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams")]
foodlevedat11s <- ndns_rp_yr11a_foodleveldietarydata_uk_20210831[,c("seriali","SubFoodGroupCode","RecipeSubFoodGroupCode","TotalGrams")]

#foodlevedat9s[ ,seriali  := factor(seriali)]
#foodlevedat10s[ ,seriali  := factor(seriali)]
#foodlevedat11s[ ,seriali  := factor(seriali)]


summary(foodlevedat9s)
summary(foodlevedat10s)
summary(foodlevedat11s)
```

```{r}
#combine with key for Nova Group

foodlevedat9s<- foodlevedat9s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat10s<- foodlevedat10s[UPFNOVA, on= "SubFoodGroupCode"]
foodlevedat11s<- foodlevedat11s[UPFNOVA, on= "SubFoodGroupCode"]



foodlevedat9s<- foodlevedat9s[,c("seriali","TotalGrams","Nova group")]
foodlevedat10s<- foodlevedat10s[,c("seriali","TotalGrams","Nova group")]
foodlevedat11s<- foodlevedat11s[,c("seriali","TotalGrams","Nova group")]

#identify total intake in grams by nova group
weights9 <-foodlevedat9s[, sum(TotalGrams), by=`Nova group`]
weights10 <-foodlevedat10s[, sum(TotalGrams), by=`Nova group`]
weights11<-foodlevedat11s[, sum(TotalGrams), by=`Nova group`]


summary(foodlevedat9s)
summary(foodlevedat10s)
summary(foodlevedat11s)


#weights11
```

```{r}

foodlevedat9s[plot(`Nova group`)]
foodlevedat10s[plot(`Nova group`)]
foodlevedat11s[plot(`Nova group`)]

plot(weights9, ylab = "Grams")
#foodlevedat9s[plot(`Nova group`, TotalGrams)]

plot(weights10, ylab = "Grams")
#foodlevedat10s[plot(`Nova group`, TotalGrams)]

plot(weights11, ylab = "Grams")
#foodlevedat11s[plot(`Nova group`, TotalGrams)]
```

```{r}
Nova11 <- xtabs(~foodlevedat11s[,`Nova group`])
Nova4 <- xtabs(~foodlevedat4s[,`Nova group`])
chi411<-rbind(Nova4,Nova11)
chisq.test(chi411)



```





```{r}
individual_intake <- foodlevedat4s[, TotalGrams, by= .(seriali,`Nova group`)]
individual_intake[ ,seriali  := factor(seriali)]
summary(individual_intake)

individual_intake[plot(TotalGrams,seriali)]

```

## how many grams from each nova group for each person? or subgroup nova 4 grams per person?

## First I write the function to process the individual entries. This totals the intake in grams and identifies the proportion by nova type.

```{r}




IndividDiet <- function(DT,individ) {
  pers <- DT[seriali==individ]
pers
oneweight<-pers[, round(sum(TotalGrams),2), by=`Nova group`][order(`Nova group`)]

setnames( oneweight, 'V1','Totalg')
 oneweight[,pcnt := round((Totalg/sum(Totalg)*100),2)]

  return(na.exclude(oneweight))
}


```

### This section uses the function designed above to process the table

```{r}

name <-foodlevedat9s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)  
#nameu
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat9s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat9sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
foodleveldat9sp



```

```{r}

name <-foodlevedat10s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat10s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat10sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
foodleveldat10sp


```


```{r}

name <-foodlevedat11s[,.(seriali)]  
name <- as.numeric(unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat11s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat11sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
foodleveldat11sp

foodleveldat910sp <- rbind(foodleveldat9sp,foodleveldat10sp)
foodleveldat911sp <- rbind(foodleveldat11sp, foodleveldat910sp)
```


```{r}

name <-foodlevedat4s[,.(seriali)]  
name <- (unlist(name))
nameu <-c(unique(name))
x <- length(nameu)
#summary(nameu)
outcome <- data.table()

for (i in 1:x){
  


output<- IndividDiet(foodlevedat4s,nameu[i])  

output$seriali <- nameu[i]

 outcome <-rbind(outcome,output) 
 
 
}

foodleveldat4sp <-dcast.data.table(outcome, seriali  ~ `Nova group` , value.var = c('Totalg',"pcnt"))
foodleveldat4sp



```


This data con now be joined with the other data

```{r}
# combine with data from version10a.rmd

setkey(sav11rp,seriali)
setkey(foodleveldat911sp, seriali)

novasav11 <-sav11rp[foodleveldat911sp, by= seriali]

setkey(sav4rp,seriali)
setkey(foodleveldat4sp,seriali)

novasav4 <-sav4rp[foodleveldat4sp, by= seriali]

```


```{r}

summary(novasav11$pcnt_4)
novasav11$pcnt_4<- as.double(novasav11$pcnt_4)
#novasav11[hist( sqrt(pcnt_4))]


summary(novasav4$pcnt_4)
novasav4$pcnt_4<- as.double(novasav4$pcnt_4)
novasav4[hist( sqrt(pcnt_4))]


ttestpcnt <- t.test(novasav4$pcnt_4,novasav11$pcnt_4)
ttestpcnt

```

```{r}
novalm11 <- lm(omsysval ~ (sqrt(pcnt_4)) + sqrt(AgeR) +Sex +Sodiummg + TotalEMJ +ethgrp2 +VitaminDµg  ,novasav11)

novalm11


novalm4 <- lm(omsysval ~ (sqrt(pcnt_4)) + sqrt(Age) +Sex + ethgr2+Sodiummg + TotalEMJ +VitaminDµg  ,novasav4)

novalm4

#novaanova <- anova(novalm11,novalm4)

```



This means that in the earlier data for each unit increase in the square root of the percentage of nova 4 food there is a corresponding increase in mean systolic bp of 0.13. That is when the intake rises from 25% to 36% meaning that  there is an increase from 5 to 6 in the square root there is an increase of 0.13 in the BP. At 100% Nova 4 there might be an increase of 1.3mmHg. 

 in 2004 increasing age from 25-36 leads to an increase in BP of 5.07 mmmHg; being female reduces the sys bp by 5.59 mmHg.




